---
title: "Simulation-Based Power Analysis"
subtitle: "Part 1.2: Between-subject design and introduction to linear modeling — Handout"
author:
  - name: "Filip Andras, David López-García & David Sánchez Casasola"
    affiliations:
      - name: "Centro de Investigación Mente Cerebro y Comportamiento (CIMCYC), Universidad de Granada"
date: "26 May 2026"
date-format: "D MMMM YYYY"
format:
  html:
    theme: cosmo
    toc: true
    toc-location: left
    toc-depth: 3
    embed-resources: true
    code-fold: false
    code-overflow: wrap
    code-tools: false
    fig-align: center
    number-sections: true
    page-layout: full
    include-in-header:
      - text: |
          <style>
            pre { white-space: pre-wrap; word-wrap: break-word; }
            .quarto-title-block { max-width: 1200px; margin: auto; padding: 0 2em; }
            main, #quarto-document-content { max-width: 1200px; margin: auto; padding: 0 2em; }
            h2 { margin-top: 2em; }
            h3 { margin-top: 1.5em; }
            #TOC, #quarto-margin-sidebar { width: 250px; }
          </style>
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| echo: false
#| eval: true

# load packages (suppress startup messages from lme4, lmerTest, etc.)
suppressPackageStartupMessages({
  library(tictoc)
  library(MASS)
  library(lme4)
  library(lmerTest)
  library(ggplot2)
})

options(scipen = 999) # prevents scientific notation
options(width = 120)  # wider console output for HTML rendering
set.seed(42)          # set seed for reproducibility
```

# Why linear models?

Part 1 introduced power analysis using independent-samples t-tests. This part
generalises the same logic to the **linear model framework**, which underlies most
modern statistical analyses in cognitive science and psycholinguistics.

Linear models are worth learning because they are a flexible framework that goes
beyond simple comparisons of means. In particular:

- They handle factorial designs and continuous predictors with the same syntax.
- They extend naturally to **linear mixed-effects models (LMEMs)** for
  repeated-measures and hierarchical data.
- They handle unbalanced data and missing observations without resorting to
  listwise deletion.
- They extend again to **generalised linear models (GLMs)** for non-Gaussian
  outcomes (accuracy, counts, etc.).
- The transition to Bayesian modeling is more straightforward when you already
  think in terms of linear models.

But the most important reason for this course is pedagogical. Linear models let us
walk through a real-life power analysis example where **analytical solutions are
not available** and where standard software like G*Power falls short.

A lot of courses end with "for a more complex design use simulations" but never
show you how. This course closes that gap.

The flip side of this flexibility is that **the data-generating process is highly
specific to your design**. There is no universal recipe — you have to know your
data well enough to specify how it is generated.

That responsibility is also the main intellectual payoff. By the time you finish
a simulation-based power analysis you understand your design better than any
analytical shortcut would have told you.

# Basic terminology

A simple linear model takes the form:

$$
y = \beta_0 + \beta_1 \cdot x + \epsilon
$$

where:

- **$y$** is the response (dependent) variable.
- **$\beta_0$** is the intercept (*intercepto*).
- **$\beta_1$** is the slope (*pendiente*).
- **$x$** is the predictor (independent variable).
- **$\epsilon$** is the residual error, $\epsilon \sim \text{Normal}(0, \sigma)$.

## Mapping to our LDT example

In the lexical-decision-task example from Part 1, the parameters map as follows:

- $\beta_0$ (intercept) = mean response time across both conditions = **725 ms**.
- $\beta_1$ (slope) = difference between conditions = **50 ms**.
- $x$ = condition (lexical frequency), coded as -0.5 / +0.5.

The predicted response times are:

- High lexical frequency: $725 + 50 \cdot (-0.5) = 700$ ms.
- Low lexical frequency: $725 + 50 \cdot (+0.5) = 750$ ms.

In this course we always code the condition as **-0.5 / +0.5** (called *effects
coding*). The intercept is the grand mean across both groups and the slope is the
**full** difference between conditions. This is convenient for power analysis: the
slope parameter directly equals the effect size you are trying to detect, with no
factor-of-2 adjustment.

R has other coding schemes (treatment 0/1, sum-to-zero -1/+1) that change the
interpretation of intercept and slope, but they all describe the same data and give
the same inferential conclusions. We stick with -0.5 / +0.5 throughout.

```{r}
#| label: redefine-betwsubj-simdat-ldt
#| echo: false
#| eval: true

# Re-define the data simulation function from Part 1 (uses -0.5 / +0.5 effects coding)
betwsubj_simdat_ldt <- function(n = 10,
                                mean_freq_high = 700,
                                mean_freq_low  = 750,
                                stdev     = 200) {
  freq_high <- rnorm(n, mean = mean_freq_high, sd = stdev)
  freq_low  <- rnorm(n, mean = mean_freq_low,  sd = stdev)
  subj     <- 1:(2 * n)
  cond     <- factor(rep(c("freq_high", "freq_low"), each = n))
  cond.cod <- ifelse(cond == "freq_high", -0.5, 0.5)
  sim_dat  <- data.frame(
    subj     = subj,
    cond     = cond,
    cond.cod = cond.cod,
    rt       = c(freq_high, freq_low)
  )
}
```

# Fitting the linear model

With the function from Part 1, we generate one simulated dataset and fit the linear
model:

```{r}
#| label: lm-summary-tval
#| echo: true
#| eval: true

sim_dat <- betwsubj_simdat_ldt()
m0 <- lm(rt ~ cond.cod, sim_dat)
summary(m0)
```

In the summary, the `(Intercept)` row is the estimate of $\beta_0$ (grand mean) and
the `cond.cod` row is the estimate of $\beta_1$ (the condition effect). With n = 10
per group, both estimates are noisy — we will see this in the parameter-recovery
section below.

# Power analysis with the linear model

The simulation logic from Part 1 works identically — only the analysis step
changes. The skeleton:

```{r}
#| label: power-loop-skeleton
#| echo: true
#| eval: true

nsim  <- 10000          # number of simulations
pvals <- rep(NA, nsim)  # vector to store p-values

for (i in 1:nsim) {
                                          # 1. generate a simulated dataset
                                          # 2. fit the linear model
                                          # 3. extract the p-value for the condition effect
}
                                          # 4. count the proportion of p-values less than 0.05
```

And the filled-in version:

```{r}
#| label: power-simulation-lm
#| echo: true
#| eval: true

nsim  <- 10000
pvals <- rep(NA, nsim)

for (i in 1:nsim) {
  sim_dat  <- betwsubj_simdat_ldt()          # 1. simulate a dataset
  m0       <- lm(rt ~ cond.cod, sim_dat)     # 2. fit the linear model
  pvals[i] <- summary(m0)$coefficients[2, 4] # 3. extract the p-value for the slope
}

round(mean(pvals < 0.05), 3)                 # 4. proportion of p < 0.05
```

The estimated power should match the t-test result from Part 1 (around 7% for n = 10
per group with a 50 ms effect and SD = 200 ms), because the t-test and linear-model
test are mathematically equivalent.

# Parameter recovery

A separate question, complementary to power, is whether the model can in principle
**recover the true parameter values** under repeated sampling. This is called
*parameter recovery*. If the model cannot recover the parameters even with infinite
data, the power analysis is moot.

```{r}
#| label: parameter-recovery-lm
#| echo: true
#| eval: true

nsim   <- 1000
params <- matrix(rep(NA, nsim * 3), ncol = 3) # 3 parameters: intercept, slope, residual SD

for (i in 1:nsim) {
  sim_dat        <- betwsubj_simdat_ldt()
  m              <- lm(rt ~ cond.cod, sim_dat)
  params[i, 1:2] <- summary(m)$coefficients[, 1] # intercept and slope
  params[i, 3]   <- summary(m)$sigma             # residual SD
}
```

We can visualise the distribution of recovered parameters and overlay the true values
(in red) used in the data-generating process:

```{r}
#| label: fig-parameter-recovery-hist
#| echo: false
#| eval: true
#| fig-cap: "Parameter recovery for the simple linear model with n = 10 per group, 1000 simulations. Red lines show the true values used to generate the data."

op <- par(mfrow = c(1, 3), pty = "s")
hist(params[, 1], main = "b0 (intercept)", col = "lightblue")
abline(v = 725, col = "red", lwd = 2)
hist(params[, 2], main = "b1 (slope)", col = "lightblue")
abline(v = 50, col = "red", lwd = 2)
hist(params[, 3], main = "sd residuals", col = "lightblue")
abline(v = 200, col = "red", lwd = 2)
par(op)
```

The histograms are centred on the true values. The model gives us **accurate
estimates of the true parameters under repeated sampling**, although they are noisy
(the slope distribution is especially wide because n = 10 is small).

Raising the number of subjects would tighten the distributions — that is precisely
the parameter recovery story behind power analysis. Later in the course we will see
designs where parameter recovery fails even with large n.

# Summary and next step

So far we have learned to perform power analysis using simulations with the
independent-samples t-test and the simple linear model. We have seen that the two
approaches give equivalent results for the between-subjects design, and we have
introduced parameter recovery as a complementary diagnostic.

The next part of the course generalises this approach to **within-subjects designs**,
where each subject contributes data to both conditions and we have to model the
dependency between observations explicitly.

This is where the linear-model framework really starts to pay off. Paired t-tests
can handle the simplest case, but linear mixed-effects models scale naturally to
designs with many repeated measures, items, and random effects.

# Further reading

- Brown, V. A. (2021). An introduction to linear mixed-effects modeling in R.
  *Advances in Methods and Practices in Psychological Science*, 4(1),
  2515245920960351.

# Session information

```{r}
#| label: session-info
#| echo: false
#| eval: true
#| code-fold: true

sessionInfo()
```
