---
title: "Simulation-Based Power Analysis"
subtitle: "Part 1.2: Between-subject design and introduction to linear modeling"
author:
  - name: "Filip Andras, David López-García & David Sánchez Casasola"
    affiliations:
      - name: "Centro de Investigación Mente Cerebro y Comportamiento (CIMCYC), Universidad de Granada"
date: "26 May 2026"
date-format: "D MMMM YYYY"
format:
  revealjs:
    theme: serif
    css: ../custom.css
    slide-number: true
    chalkboard: true
    code-fold: false
    code-overflow: wrap
    scrollable: true
    footer: "Justificación del tamaño muestral y análisis de potencia estadística | v2026.2"
execute:
  warning: false
  message: false
---

## Part 1.2: Between-subject design and linear modeling

```{r}
#| label: setup
#| echo: false
#| eval: true

# load packages (suppress startup messages from lme4, lmerTest, etc.)
suppressPackageStartupMessages({
  library(tictoc)
  library(MASS)
  library(lme4)
  library(lmerTest)
  library(ggplot2)
})

options(scipen = 999) # prevents scientific notation
options(width = 120)  # wider console output for HTML rendering
set.seed(42)          # set seed for reproducibility
```

**Why linear models?**

::: {.incremental}
- Linear models (LM) are a flexible framework that goes beyond simple comparisons of means (t-tests, ANOVAs).
- Not only factorial designs, we can have (multiple) continuous predictors.
- They can be easily extended to linear mixed-effects models (LMEM) for more complex data structures (e.g., repeated measures, hierarchical data).
- ...to model variability at multiple levels (e.g., subjects, items).
- ...handles unbalanced data and missing data better (no listwise deletion).
- Easily extends to generalized linear models (GLMs) for a variety of response variables (e.g., accuracy, counts etc.).
- The transition to Bayesian modeling is more straightforward.
:::

## Part 1.2: Between-subject design and linear modeling

**Why linear models?**

::: {.incremental}
- More importantly, to walk you through a real-life example where analytical solutions are not available and where existing software (G*Power) falls short.
- A lot of courses end with "for a more complex design use simulations" but do not show you how to do it.
- Comes with a disadvantage/advantage: the data generative process (e.g., how we simulate the data) is highly specific to your design.
- How well do you know your data?
:::

## Part 1.2: Between-subject design and linear modeling

**What is a linear model? (basic terminology)**

$y  = \beta_0 + \beta_1 * x + \epsilon$

- **Where:**
- y = response/dependent variable
- $\beta_0$ = intercept = intercepto
- $\beta_1$ = slope = pendiente
- x = predictor/independent variable
- $\epsilon$ = residual error ~ Normal(0, σ)

## Part 1.2: Between-subject design and linear modeling

**In our example:**

$y  = \beta_0 + \beta_1 * x + \epsilon$

::: {.incremental}
- $\beta_0$ (intercept) = mean response times across both conditions = 725 ms
- $\beta_1$ (slope) = the difference between the two conditions = 50 ms
- x = condition = lexical frequency (coded as -0.5 / +0.5)
:::

## Part 1.2: Between-subject design and linear modeling

**In our example:**

$y  = \beta_0 + \beta_1 * x + \epsilon$

- $\beta_0$ (intercept) = mean response times across both conditions = 725 ms
- $\beta_1$ (slope) = the difference between the two conditions = 50 ms
- x = condition = lexical frequency (coded as -0.5 / +0.5)

**Response times for words with high lexical frequency:**

- 725 + 50*(-0.5) = 700

**Response times for words with low lexical frequency:**

- 725 + 50*(+0.5) = 750

**-0.5 and +0.5 coding is called "effects coding", which is not the default**

## Part 1.2: Between-subject design and linear modeling

```{r}
#| label: redefine-betwsubj-simdat-ldt
#| echo: false
#| eval: true

# Specify the data simulation function that we defined earlier:
betwsubj_simdat_ldt <- function(n = 10,
                                mean_freq_high = 700,
                                mean_freq_low = 750,
                                stdev = 200) {
  freq_high <- rnorm(n, mean = mean_freq_high, sd = stdev)
  freq_low  <- rnorm(n, mean = mean_freq_low,  sd = stdev)
  subj <- 1:(2 * n)
  cond <- factor(rep(c("freq_high", "freq_low"), each = n))
  cond.cod <- ifelse(cond == "freq_high", -0.5, 0.5)
  # put it together:
  sim_dat <- data.frame(
    subj = subj,
    cond = cond,
    cond.cod = cond.cod,
    rt = c(freq_high, freq_low)
  )
}
```

**A note on coding the condition variable**

In this course we always code the two-level condition as **-0.5 / +0.5** (called *effects coding*).

::: {.incremental}
- The intercept ($\beta_0$) is the grand mean across both groups.
- The slope ($\beta_1$) is the **full** difference between conditions.
- Predicted RT for high frequency: $725 + 50 \cdot (-0.5) = 700$ ms.
- Predicted RT for low frequency: $725 + 50 \cdot (+0.5) = 750$ ms.
:::

R has other coding schemes (treatment 0/1, sum-to-zero -1/+1) but effects coding is the most convenient for power analysis: the slope parameter equals the effect size you are trying to detect. No factor-of-2 adjustment.

## Part 1.2: Between-subject design and linear modeling

**Generate one simulated dataset and fit a linear model:**

```{r}
#| label: lm-summary-tval
#| echo: true
#| eval: true
#| output-location: fragment

sim_dat <- betwsubj_simdat_ldt()
m0 <- lm(rt ~ cond.cod, sim_dat)
summary(m0)
```

## Part 1.2: Between-subject design and linear modeling

**Power analysis through simulation with a simple linear model**

**The same logic as with the t-test applies:**

```{r}
#| label: power-loop-skeleton
#| echo: true
#| eval: true

nsim <- 10000          # determine the number of simulations
pvals <- rep(NA, nsim) # create an empty vector to store p-values

for(i in 1:nsim) {
                                          # 1. generate a simulated dataset
                                          # 2. fit the model
                                          # 3. extract the p-value for the condition effect
}
                                          # 4. count the proportion of p-values less than 0.05
```

## Part 1.2: Between-subject design and linear modeling

**Power analysis through simulation with a simple linear model**

**The same logic as with the t-test applies:**

```{r}
#| label: power-simulation-lm
#| echo: true
#| eval: true
#| output-location: fragment

nsim <- 10000          # determine the number of simulations
pvals <- rep(NA, nsim) # create an empty vector to store p-values

for(i in 1:nsim) {
  sim_dat <- betwsubj_simdat_ldt()          # 1. simulate a dataset
  m0 <- lm(rt ~ cond.cod, sim_dat)          # 2. fit the model
  pvals[i] <- summary(m0)$coefficients[2,4] # 3. extract the p-value for the condition effect
}
round(mean(pvals < 0.05), 3)                # 4. count the proportion of p-values less than 0.05
```

## Part 1.2: Between-subject design and linear modeling

**Parameter Recovery**

Can the model, in principle, recover the true parameters under repeated sampling?

```{r}
#| label: parameter-recovery-lm
#| echo: true
#| eval: true
#| code-line-numbers: "||1|3-5|8|9|10|11|1-12"

nsim <- 1000

# create empty matrix to store parameters, there are only three: intercept, slope
## and the standard deviation
params <- matrix(rep(NA, nsim*3), ncol = 3)

for(i in 1:nsim) {
  sim_dat <- betwsubj_simdat_ldt()
  m <- lm(rt ~ cond.cod, sim_dat)
  params[i,c(1,2)] <- summary(m)$coefficients[,1]
  params[i,3] <- summary(m)$sigma
}
```

## Part 1.2: Between-subject design and linear modeling

**Parameter Recovery Visualization:**

```{r}
#| label: fig-parameter-recovery-hist
#| echo: false
#| eval: true

# plot the distribution of the parameters under repeated sampling
params_df <- data.frame(
  intercept = params[,1],
  slope = params[,2],
  sd = params[,3]
)

# overlay with true parameters (red lines show true values)
op <- par(mfrow = c(1,3), pty = "s")
hist(params[,1], main = "b0", col = "lightblue")
abline(v=725, col = "red", lwd = 2)
hist(params[,2], main = "b1", col = "lightblue")
abline(v=50, col = "red", lwd = 2)
hist(params[,3], main = "sd residuals", col = "lightblue")
abline(v=200, col = "red", lwd = 2)
par(op)
```
:::{.incremental}
- Looks ok. 
- It tells us that the model can give us accurate estimates of the true parameters under repeated sampling. If we raise the number of subjects, we will have more precise estimates.
- Later we will see when parameter recovery fails.
:::

## In this section we have seen:

::: {.incremental}
- How to perform power analysis using simulations with independent samples t-tests and linear models.
- We will now simulate data for a within-subjects design.
:::

## Part 1.2: Between-subject design and linear modeling

**Gentle introduction to linear (mixed effects) models:**

Brown, V. A. (2021). An introduction to linear mixed-effects modeling in R. Advances in Methods and Practices in Psychological Science, 4(1), 2515245920960351.

##  Preguntas? {background-color="#43464B"}

## Session information {.smaller}

```{r}
#| label: session-info
#| echo: false
#| eval: true

sessionInfo()
```
