---
title: "Simulation-Based Power Analysis"
subtitle: "Part 3: Repeated measures — Handout"
author:
  - name: "Filip Andras, David López-García & David Sánchez Casasola"
    affiliations:
      - name: "Centro de Investigación Mente Cerebro y Comportamiento (CIMCYC), Universidad de Granada"
date: "26 May 2026"
date-format: "D MMMM YYYY"
format:
  html:
    theme: cosmo
    toc: true
    toc-location: left
    toc-depth: 3
    embed-resources: true
    code-fold: false
    code-overflow: wrap
    code-tools: false
    fig-align: center
    number-sections: true
    page-layout: full
    include-in-header:
      - text: |
          <style>
            pre { white-space: pre-wrap; word-wrap: break-word; }
            .quarto-title-block { max-width: 1200px; margin: auto; padding: 0 2em; }
            main, #quarto-document-content { max-width: 1200px; margin: auto; padding: 0 2em; }
            h2 { margin-top: 2em; }
            h3 { margin-top: 1.5em; }
            #TOC, #quarto-margin-sidebar { width: 250px; }
          </style>
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| echo: false
#| eval: true

# load packages (suppress startup messages from lme4, lmerTest, etc.)
suppressPackageStartupMessages({
  library(tictoc)
  library(MASS)
  library(lme4)
  library(lmerTest)
  library(ggplot2)
})

options(scipen = 999) # prevents scientific notation
options(width = 120)  # wider console output for HTML rendering
set.seed(42)          # set seed for reproducibility
```

# From one observation per condition to many

Part 2 simulated within-subjects data with **one observation per subject per
condition**: each person contributed two RTs, one high-frequency and one
low-frequency.

In reality, cognitive experiments collect many trials per condition per subject.
A typical lexical-decision task might include 20 high-frequency words and 20
low-frequency words per participant.

This part of the course generalises the data-generating process to **multiple
repeated measures per condition per subject**. The random-intercept idea from
Part 2 carries over directly; we just need to be careful with how we replicate
the subject-level structure.

# Building the design matrix

For 20 subjects $\times$ 20 trials per condition $\times$ 2 conditions, the data
frame has $20 \times 20 \times 2 = 800$ rows. The condition vector has to be made
$n \times k$ times long for each condition. The subject vector has to repeat each
subject id $k$ times for one condition, and then the whole vector has to be
repeated twice for the two conditions.

```{r}
#| label: build-repeated-measures-design
#| echo: true
#| eval: true

n <- 20  # subjects
k <- 20  # 20 observations per condition (40 trials per subject in total)

# condition vector: each condition repeated n * k times
cond.cod <- rep(c(-0.5, 0.5), each = n * k)

# subject vector: each of the n subject ids repeated k times,
# then the whole vector repeated twice (once per condition):
subj <- rep(rep(1:n, each = k), times = 2)

# put it together:
sim_dat <- data.frame(
  subj     = subj,
  cond.cod = cond.cod
)
```

Sanity checks:

```{r}
#| label: check-dim-sim-dat
#| echo: true
#| eval: true

dim(sim_dat) # should be 800 x 2
```

The cross-tabulation confirms that every subject has 20 observations in each of
the two conditions:

```{r}
#| label: xtabs-subj-by-cond
#| echo: true
#| eval: true

xtabs(~ subj + cond.cod, sim_dat)
```

# Generating the response times

To extend the Part 2 generative formula to multiple trials per condition, the
by-subject intercept adjustment ($u_{0i}$) has to be **repeated $k$ times** for
each subject and then twice (once per condition):

```{r}
#| label: generate-rt-repeated-measures
#| echo: true
#| eval: true

# previously (Part 2: one trial per condition per subject):
subject_adj <- round(rnorm(n, mean = 0, sd = 150))
# rt <- 725 + rep(subject_adj, times = 2) + 50 * cond.cod + rnorm(2 * n, 0, 200)

# now (Part 3: k trials per condition per subject):
sim_dat$rt <- 725 +
              rep(rep(round(rnorm(n, 0, 150)), each = k), times = 2) + # subject adjustments
              50 * cond.cod +                                          # condition effect
              rnorm(n * k * 2, 0, 200)                                 # residual error
```

The nested `rep(rep(..., each = k), times = 2)` is the key change:

- The inner `rep(..., each = k)` says "each subject's adjustment is repeated $k$
  times" (one value per trial).
- The outer `rep(..., times = 2)` says "the whole pattern is then repeated twice
  for the two conditions".

Without this nested structure the subject adjustments would not be aligned with
the right rows of `sim_dat`.

# Fitting the random-intercept model

With multiple trials per condition per subject, the paired t-test is no longer
applicable. We need the linear mixed-effects model with a by-subject random
intercept:

```{r}
#| label: lmer-varying-intercepts-m2
#| echo: true
#| eval: true

m2 <- lmer(rt ~ cond.cod + (1 | subj), sim_dat,
           control = lmerControl(calc.derivs = FALSE))
summary(m2)
```

The `(1 | subj)` term has the same meaning as in Part 2: each subject has their
own intercept drawn from a normal distribution centred on the grand mean. The
only difference is that now each subject contributes $2k = 40$ trials instead of
just 2.

# A first look at varying slopes

If we suspect that the lexical-frequency effect itself varies across subjects —
that is, some people show a bigger effect than others — we can fit a model that
allows **both** the intercept and the slope to vary by subject:

```{r}
#| label: lmer-varying-slopes-m3
#| echo: true
#| eval: true

m3 <- lmer(rt ~ cond.cod + (1 + cond.cod | subj), sim_dat,
           control = lmerControl(calc.derivs = FALSE))
summary(m3)
```

The new `(1 + cond.cod | subj)` term says "each subject has their own intercept
*and* their own slope, correlated with each other".

::: callout-warning
**Look at the correlation between intercept and slope in the random-effects
section.** If you see a correlation of -1 or +1, the model is **overparameterized**
— there is not enough information in the data to estimate the correlation
properly.

This is exactly what we see here. The reason is that we generated the data
**without** by-subject slope variability: the slope was a constant 50 ms for every
subject. The model is trying to estimate a correlation between intercept and a
slope that does not vary, which is mathematically ill-defined.
:::

The fix is to add by-subject slope variability to the data-generating process,
which is the topic of Part 4.

# Summary and next step

So far in this part we have:

- Extended the within-subjects design to **multiple trials per condition per
  subject**, the realistic case in cognitive experiments.
- Built the design matrix with nested `rep()` calls so that each subject's
  intercept adjustment aligns with their $2k$ trials.
- Fit the by-subject random-intercept LMM with `lmer(rt ~ cond.cod + (1 | subj))`.
- Tried a varying-slopes model and saw a $\pm 1$ intercept-slope correlation —
  the diagnostic signal that our generative process is missing slope variability.

The next part of the course **adds by-subject adjustments to the slope** to the
data-generating process, so the model is no longer trying to estimate a phantom
correlation.

# Session information

```{r}
#| label: session-info
#| echo: false
#| eval: true
#| code-fold: true

sessionInfo()
```
