---
title: "Simulation-Based Power Analysis"
subtitle: "Part 6: Fully crossed design and by-item varying intercepts — Handout"
author:
  - name: "Filip Andras, David López-García & David Sánchez Casasola"
    affiliations:
      - name: "Centro de Investigación Mente Cerebro y Comportamiento (CIMCYC), Universidad de Granada"
date: "26 May 2026"
date-format: "D MMMM YYYY"
format:
  html:
    theme: cosmo
    toc: true
    toc-location: left
    toc-depth: 3
    embed-resources: true
    code-fold: false
    code-overflow: wrap
    code-tools: false
    fig-align: center
    number-sections: true
    page-layout: full
    include-in-header:
      - text: |
          <style>
            pre { white-space: pre-wrap; word-wrap: break-word; }
            .quarto-title-block { max-width: 1200px; margin: auto; padding: 0 2em; }
            main, #quarto-document-content { max-width: 1200px; margin: auto; padding: 0 2em; }
            h2 { margin-top: 2em; }
            h3 { margin-top: 1.5em; }
            #TOC, #quarto-margin-sidebar { width: 250px; }
          </style>
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| echo: false
#| eval: true

# load packages (suppress startup messages from lme4, lmerTest, etc.)
suppressPackageStartupMessages({
  library(tictoc)
  library(MASS)
  library(lme4)
  library(lmerTest)
  library(ggplot2)
})

options(scipen = 999) # prevents scientific notation
options(width = 120)  # wider console output for HTML rendering
set.seed(42)          # set seed for reproducibility
```

# Why by-item random intercepts?

So far we have modelled variability between **subjects**: some people are
faster, some slower, some show a bigger lexical-frequency effect, others a
smaller one.

But the **items** (the actual words in our LDT example) also come with their
own variability. Words are sampled from a pool of possible items. Each word has
its own baseline RT: *table* might be responded to consistently faster than
*xylophone*, regardless of who is reading.

To make the simulation realistic, we add **by-item adjustments to the
intercept**.

# Fully crossed designs

A second important structural feature: in this example each word is seen by
**every subject**. This is called a **fully crossed design**. Every (subject,
item) cell of the design matrix has at least one observation.

Two consequences:

- The data are not independent in two different ways: responses to the same
  item are correlated (across subjects), and responses from the same subject
  are correlated (across items).
- A traditional analysis would have run two ANOVAs — one averaged over
  subjects (called $F_1$) and one averaged over items ($F_2$). Linear mixed
  models replace both with a single model that includes both sources of
  variability.

The LMM formula notation for the item random intercept is `(1 | item)`, exactly
parallel to `(1 | subj)` for subjects.

# Building the crossed design

We use `expand.grid()` to generate every combination of `item` and `subj`:

```{r}
#| label: build-crossed-design
#| echo: true
#| eval: true

n <- 20  # subjects
k <- 20  # 20 items per condition (so 40 items total)

# fully crossed design: each subject sees all items
sim_dat <- expand.grid(item = 1:(k * 2), subj = 1:n)

# assign items to conditions (first k items are high frequency, next k are low)
sim_dat$cond.cod <- rep(c(-0.5, 0.5), each = k)

head(sim_dat)
```

Sanity check — every subject sees every item exactly once:

```{r}
#| label: xtabs-subj-by-item
#| echo: true
#| eval: true

head(xtabs(~ subj + item, data = sim_dat))
```

Each cell contains a 1, confirming the fully crossed structure.

# Generating by-item adjustments

By-item adjustments are drawn from a normal distribution centred on zero, the
same way we did for subjects. A reasonable starting value is $\sigma_w = 100$:

```{r}
#| label: generate-item-intercepts-w
#| echo: true
#| eval: true

sigma_w <- 100   # SD of item random intercepts

# one adjustment per unique item:
w <- rnorm(n = length(unique(sim_dat$item)), mean = 0, sd = sigma_w)
w
```

There is one value per item. We will look it up by item id during data
generation, the same way we looked up subject adjustments in Part 4.

# Extended generative formula

The full generative formula now includes both a subject and an item random
intercept:

$$
rt_{ij} = \beta_0 + u_{0i} + w_{0j} + (\beta_1 + u_{1i}) \times \text{cond.cod}_{ij} + \varepsilon_{ij}
$$

with:

- $u_{0i} \sim \text{Normal}(0, \sigma_{u0})$ — by-subject intercept
- $u_{1i} \sim \text{Normal}(0, \sigma_{u1})$ — by-subject slope (correlated with $u_{0i}$)
- $w_{0j} \sim \text{Normal}(0, \sigma_w)$ — by-item intercept (the new term)
- $\varepsilon_{ij} \sim \text{Normal}(0, \sigma)$ — residual error

The model fits this with: `rt ~ cond.cod + (1 + cond.cod | subj) + (1 | item)`.

# A function for the full crossed design

We extend Part 5's generator one more time with the item random intercept:

```{r}
#| label: define-gendat-ldt-crossed
#| echo: true
#| eval: true

gendat_ldt_crossed <- function(n        = 20,    # subjects
                               k        = 20,    # items per condition
                               b0       = 725,   # grand mean
                               b1       = 50,    # frequency effect
                               sigma_u0 = 150,   # SD of subject intercepts
                               sigma_u1 = 20,    # SD of subject slopes
                               rho      = -0.5,  # correlation between subject intercepts and slopes
                               sigma_w  = 100,   # SD of item intercepts (new)
                               sigma    = 200) { # residual SD

  # fully crossed design
  sim_dat          <- expand.grid(item = 1:(k * 2), subj = 1:n)
  sim_dat$cond.cod <- rep(c(-0.5, 0.5), each = k)

  # subject random effects: correlated intercepts and slopes
  Sigma_u <- matrix(c(sigma_u0^2,                  rho * sigma_u0 * sigma_u1,
                      rho * sigma_u0 * sigma_u1,   sigma_u1^2),
                    nrow = 2)
  u <- mvrnorm(n     = length(unique(sim_dat$subj)),
               mu    = c(0, 0),
               Sigma = Sigma_u)

  # item random effects: intercepts only
  w <- rnorm(n = length(unique(sim_dat$item)), mean = 0, sd = sigma_w)

  # generate RTs row by row
  N  <- dim(sim_dat)[1]
  rt <- rep(NA, N)
  for (i in 1:N) {
    rt[i] <- rnorm(1,
                   b0 +
                   u[sim_dat[i, ]$subj, 1] +       # subject intercept
                   w[sim_dat[i, ]$item] +          # item intercept (new)
                   (b1 + u[sim_dat[i, ]$subj, 2]) * sim_dat[i, ]$cond.cod,
                   sigma)
  }

  sim_dat$rt   <- rt
  sim_dat$subj <- factor(sim_dat$subj)
  sim_dat$item <- factor(sim_dat$item)
  sim_dat
}
```

The two new lines compared to Part 5: the `rnorm()` call that generates `w`,
and the `w[sim_dat[i, ]$item]` lookup inside the loop.

# Testing the function

Generate one dataset and inspect its structure:

```{r}
#| label: test-gendat-ldt-crossed
#| echo: true
#| eval: true

sim_dat <- gendat_ldt_crossed(n = 20, k = 20)
dim(sim_dat) # 20 subjects x 40 items = 800 observations
```

The first 41 rows show a single subject's data — 40 trials (20 per condition):

```{r}
#| label: head-sim-dat-crossed
#| echo: true
#| eval: true

head(sim_dat, 41)
```

## Verifying the fully crossed structure

Each subject sees 20 items per condition:

```{r}
#| label: verify-subj-by-cond
#| echo: true
#| eval: true

xtabs(~ subj + cond.cod, data = sim_dat)
```

Each item is responded to by all 20 subjects:

```{r}
#| label: verify-cond-by-item
#| echo: true
#| eval: true

xtabs(~ cond.cod + item, data = sim_dat)
```

Exactly one observation per (subject, item) pair:

```{r}
#| label: verify-subj-by-item
#| echo: true
#| eval: true

xtabs(~ subj + item, data = sim_dat)
```

# What do the simulated RTs look like?

Generate a larger dataset (64 subjects, 40 items per condition) and plot the
histogram:

```{r}
#| label: fig-hist-simulated-rt-normal
#| echo: true
#| eval: true
#| fig-cap: "Histogram of simulated RTs from the fully crossed design. Notice the symmetric, normal-like distribution — which is unrealistic for real RT data."

sim_dat <- gendat_ldt_crossed(n = 64, k = 40)
hist(sim_dat$rt, breaks = 30,
     main = "Histogram of simulated RTs",
     xlab = "Reaction Time (ms)")
```

::: callout-warning
**Look closely at the distribution.** It is roughly symmetric and bell-shaped.
But real reaction times are **right-skewed** — there is a hard lower bound
(motor response cannot be faster than ~150 ms) and a long right tail (occasional
very slow trials).

The Gaussian generative process produces RTs that can even go *negative* for
extreme outliers, which is impossible in reality. Part 7 fixes this by
switching to a **log-normal** distribution for the RTs.
:::

# Summary and next step

In this part we:

- Added **by-item random intercepts** as a third random effect.
- Recognised that **fully crossed designs** carry dependency from both subjects
  and items, which is naturally handled by `(1 | subj) + (1 | item)` in the LMM.
- Extended the data-generating function to `gendat_ldt_crossed()` with
  $\sigma_w$ as a new parameter.
- Verified the design is fully crossed.
- Noted that Gaussian-generated RTs are unrealistically symmetric.

The current generative process is now:

$$
rt_{ij} = \beta_0 + u_{0i} + w_{0j} + (\beta_1 + u_{1i}) \times \text{cond.cod}_{ij} + \varepsilon_{ij}
$$

with $\varepsilon_{ij} \sim \text{Normal}(0, \sigma)$,

$$
\begin{pmatrix} u_0 \\ u_1 \end{pmatrix} \sim \text{MVN} \left(
\begin{pmatrix} 0 \\ 0 \end{pmatrix}, \, \Sigma_u \right), \quad
\Sigma_u = \begin{pmatrix}
\sigma_{u0}^2 & \rho \, \sigma_{u0} \sigma_{u1} \\
\rho \, \sigma_{u0} \sigma_{u1} & \sigma_{u1}^2
\end{pmatrix},
$$

and $w_0 \sim \text{Normal}(0, \sigma_w)$.

The next part of the course models **$\log(rt)$ instead of $rt$**, which both
ensures positive reaction times and captures the right-skewed shape of
real RT distributions.

# Session information

```{r}
#| label: session-info
#| echo: false
#| eval: true
#| code-fold: true

sessionInfo()
```
