---
title: "Simulation-Based Power Analysis"
subtitle: "Part 7: Modeling reaction time data — log-normal distribution — Handout"
author:
  - name: "Filip Andras, David López-García & David Sánchez Casasola"
    affiliations:
      - name: "Centro de Investigación Mente Cerebro y Comportamiento (CIMCYC), Universidad de Granada"
date: "26 May 2026"
date-format: "D MMMM YYYY"
format:
  html:
    theme: cosmo
    toc: true
    toc-location: left
    toc-depth: 3
    embed-resources: true
    code-fold: false
    code-overflow: wrap
    code-tools: false
    fig-align: center
    number-sections: true
    page-layout: full
    include-in-header:
      - text: |
          <style>
            pre { white-space: pre-wrap; word-wrap: break-word; }
            .quarto-title-block { max-width: 1200px; margin: auto; padding: 0 2em; }
            main, #quarto-document-content { max-width: 1200px; margin: auto; padding: 0 2em; }
            h2 { margin-top: 2em; }
            h3 { margin-top: 1.5em; }
            #TOC, #quarto-margin-sidebar { width: 250px; }
          </style>
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| echo: false
#| eval: true

# load packages (suppress startup messages from lme4, lmerTest, etc.)
suppressPackageStartupMessages({
  library(tictoc)
  library(MASS)
  library(lme4)
  library(lmerTest)
  library(ggplot2)
})

options(scipen = 999) # prevents scientific notation
options(width = 120)  # wider console output for HTML rendering
set.seed(42)          # set seed for reproducibility
```

# The shape of real reaction times

Part 6 left us with a Gaussian-distributed RT that is symmetric and bell-shaped.
Real reaction times are not like that. They are **right-skewed**: there is a
hard physical lower bound (motor response cannot be faster than ~150-200 ms) and
a long tail of slow trials.

A common way to model right-skewed positive data is the **log-normal
distribution**: if $\log(\text{RT})$ is normally distributed, then RT itself is
log-normal and strictly positive.

This part of the course:

- Compares Gaussian and log-normal models on **real LDT data**.
- Builds a generator that produces log-normal RTs.
- Discusses how to **back-transform** effects from the log scale to the
  millisecond scale.
- Runs a power analysis and a Type-I error check.

# Real LDT data

Load a real LDT dataset (the same one used in Filip's PhD work):

```{r}
#| label: load-ldt-data
#| echo: true
#| eval: true

library(rio)
dat <- import("../data/raw_data/ldt_data.csv")
head(dat)
```

The dataset has one row per trial, with columns for subject, stimulus,
condition (`wrd_freq.cod`), and RT in milliseconds.

# Fitting the model: Gaussian vs log-normal

We compare two models on the same data: one assumes RT is normally
distributed, the other models $\log(\text{RT})$. Click between the tabs to
compare the model output and the diagnostic panels.

::: panel-tabset

## Gaussian (raw RT)

```{r}
#| label: lmer-normal-rt
#| echo: true
#| eval: true

m <- lmer(RT ~ wrd_freq.cod + (1 + wrd_freq.cod | subject) + (1 | stimuli),
          data = dat)
summary(m)
```

`performance::check_model()` produces a diagnostic panel for the model
assumptions:

```{r}
#| label: fig-check-model-normal
#| echo: true
#| eval: true
#| fig-width: 9
#| fig-height: 9
#| fig-cap: "Model diagnostics for RT modeled as Gaussian. Note the strong departure from normality in the QQ plot and the heteroscedasticity in the residuals-vs-fitted plot."

performance::check_model(m)
```

The QQ plot of residuals shows strong departure from normality in the upper
tail, and the residuals-vs-fitted plot shows heteroscedasticity (the variance
increases with the fitted value). Both are typical signatures of right-skewed
data.

## Log-normal (log(RT))

```{r}
#| label: lmer-log-rt
#| echo: true
#| eval: true

m_log <- lmer(log(RT) ~ wrd_freq.cod + (1 + wrd_freq.cod | subject) + (1 | stimuli),
              data = dat)
summary(m_log)
```

Same diagnostic panel for the log-transformed model:

```{r}
#| label: fig-check-model-log
#| echo: true
#| eval: true
#| fig-width: 9
#| fig-height: 9
#| fig-cap: "Model diagnostics for log(RT). The QQ plot is much closer to the diagonal and the residual variance is more stable across fitted values."

performance::check_model(m_log)
```

The QQ plot is now much closer to the diagonal and the residual variance is
more uniform. This is a substantial improvement.

:::

The log-normal model wins on both diagnostic criteria. From here on the
course commits to modelling $\log(\text{RT})$ and the simulation generator
switches from `rnorm()` to `rlnorm()`.

# Back to the millisecond scale

Fitting on the log scale costs us interpretability: the coefficients are no
longer in milliseconds. To recover ms-scale effects we back-transform with
`exp()`.

With the fitted coefficients $b_0 = 6.59$, $b_1 = 0.10$ and $\text{SE}(b_1) =
0.02$:

```{r}
#| label: backtransform-all
#| echo: true
#| eval: true
#| code-fold: true
#| code-summary: "Show code (back-transform + 95% CI)"

b0        <- 6.59
b1        <- 0.10
b1_stderr <- 0.02

# point estimate
rt_high <- exp(b0 + b1 * -0.5)
rt_low  <- exp(b0 + b1 *  0.5)
effect  <- rt_low - rt_high

# 95% CI on b1 (log scale), then back-transform
lower <- b1 - 1.96 * b1_stderr
upper <- b1 + 1.96 * b1_stderr

effect_lower <- exp(b0 + lower * 0.5) - exp(b0 + lower * -0.5)
effect_upper <- exp(b0 + upper * 0.5) - exp(b0 + upper * -0.5)

c(rt_high      = rt_high,
  rt_low       = rt_low,
  effect_ms    = effect,
  effect_lower = effect_lower,
  effect_upper = effect_upper)
```

Two takeaways:

- The point estimate is a **~70 ms effect**, but this is conditional on the
  baseline RT being around 730 ms. The same log-scale effect ($b_1 = 0.10$)
  implies a different ms-scale effect at different baselines. **This is the
  most important practical consequence of log-scale modelling.**
- The 95% CI on the ms scale spans roughly **40 to 100 ms** — substantial
  uncertainty, worth keeping in mind when interpreting the point estimate.

::: callout-tip
In Bayesian terms, $b_1 \sim \text{Normal}(0.10, 0.02)$. The back-transformed
range above corresponds to a 95% credible interval on the ms-scale effect.
:::

# Generating log-normal RTs

To run a power analysis on this model, we need a data generator that produces
**log-normal** RTs. The only changes from Part 6's `gendat_ldt_crossed()`:

- All parameters are now on the **log scale** (so smaller numbers).
- `rnorm()` becomes `rlnorm()` inside the row-by-row loop.

```{r}
#| label: define-gendat-ldt-crossed-log
#| echo: true
#| eval: true

gendat_ldt_crossed_log <- function(n        = 20,    # subjects
                                   k        = 20,    # items per condition
                                   b0       = 6.59,  # log(grand mean RT); exp(6.59) ~ 730 ms
                                   b1       = 0.10,  # log-scale frequency effect (~70 ms on RT scale)
                                   sigma_u0 = 0.2,   # SD of subject intercepts (log)
                                   sigma_u1 = 0.02,  # SD of subject slopes (log)
                                   rho      = -0.5,  # correlation between subject intercepts and slopes
                                   sigma_w  = 0.1,   # SD of item intercepts (log)
                                   sigma    = 0.25) {# residual SD (log)

  # fully crossed design
  sim_dat          <- expand.grid(item = 1:(k * 2), subj = 1:n)
  sim_dat$cond.cod <- rep(c(-0.5, 0.5), each = k)

  # subject random effects (correlated intercepts and slopes)
  Sigma_u <- matrix(c(sigma_u0^2,                  rho * sigma_u0 * sigma_u1,
                      rho * sigma_u0 * sigma_u1,   sigma_u1^2),
                    nrow = 2)
  u <- mvrnorm(n     = length(unique(sim_dat$subj)),
               mu    = c(0, 0),
               Sigma = Sigma_u)

  # item random effects (intercepts only)
  w <- rnorm(n = length(unique(sim_dat$item)), mean = 0, sd = sigma_w)

  # generate RTs row by row -- note rlnorm() instead of rnorm()
  N  <- dim(sim_dat)[1]
  rt <- rep(NA, N)
  for (i in 1:N) {
    rt[i] <- rlnorm(1,
                    b0 +
                    u[sim_dat[i, ]$subj, 1] +
                    w[sim_dat[i, ]$item] +
                    (b1 + u[sim_dat[i, ]$subj, 2]) * sim_dat[i, ]$cond.cod,
                    sigma)
  }

  sim_dat$rt   <- rt
  sim_dat$subj <- factor(sim_dat$subj)
  sim_dat$item <- factor(sim_dat$item)
  sim_dat
}
```

Check that the resulting distribution is right-skewed:

```{r}
#| label: fig-hist-lognormal-rt
#| echo: true
#| eval: true
#| fig-cap: "Distribution of simulated RTs from the log-normal generator. The right-skewed shape matches what we see in real RT data."

sim_dat <- gendat_ldt_crossed_log(n = 100, k = 100)
hist(sim_dat$rt,
     main = "Distribution of Simulated RTs (Log-Normal)",
     xlab = "RT (ms)", col = "lightblue", breaks = 50)
```

Compare with the Gaussian histogram from Part 6 — the log-normal version has
the characteristic right tail and a hard zero lower bound.

# Sanity check on the log-normal model

Fit the same log(rt) model on simulated data to confirm the generator and the
model agree:

```{r}
#| label: lmer-lognormal-sanity
#| echo: true
#| eval: true

sim_dat <- gendat_ldt_crossed_log(n = 20, k = 20)

m <- lmer(log(rt) ~ cond.cod + (1 + cond.cod | subj) + (1 | item), sim_dat,
          control = lmerControl(calc.derivs = FALSE))
summary(m)
```

The fixed-effect estimates should be close to the generating values (6.59 and
0.10).

# Power analysis

With the log-normal generator in hand, the power loop is identical in shape to
the previous parts — only the simulator and model formula change:

```{r}
#| label: power-simulation-lognormal
#| echo: true
#| eval: true

nsim  <- 1000
pvals <- rep(NA, nsim)

for (i in 1:nsim) {
  sim_dat <- gendat_ldt_crossed_log(n = 20, k = 20,
                                    b0 = 6.59, b1 = 0.10,
                                    sigma_u0 = 0.2, sigma_u1 = 0.02,
                                    rho = -0.5, sigma_w = 0.1, sigma = 0.25)

  m <- lmer(log(rt) ~ cond.cod + (1 + cond.cod | subj) + (1 | item), sim_dat,
            control = lmerControl(calc.derivs = FALSE))
  pvals[i] <- summary(m)$coefficients[2, 5]
}

power <- round(mean(pvals < 0.05), digits = 4)
power
```

# Parameter recovery (seven parameters now)

With the addition of $\sigma_w$ and $\rho$ on top of the standard four
parameters, we now have **seven** parameters to recover:

```{r}
#| label: parameter-recovery-lognormal
#| echo: true
#| eval: true

nsim   <- 1000
params <- matrix(rep(NA, nsim * 7), ncol = 7)

for (i in 1:nsim) {
  sim_dat <- gendat_ldt_crossed_log(n = 20, k = 20,
                                    b0 = 6.59, b1 = 0.10,
                                    sigma_u0 = 0.2, sigma_u1 = 0.02,
                                    rho = -0.5, sigma_w = 0.1, sigma = 0.25)
  m <- lmer(log(rt) ~ cond.cod + (1 + cond.cod | subj) + (1 | item), sim_dat,
            control = lmerControl(calc.derivs = FALSE))

  params[i, 1] <- summary(m)$coefficients[1, 1] # b0
  params[i, 2] <- summary(m)$coefficients[2, 1] # b1
  params[i, 3] <- attr(VarCorr(m), "sc")        # residual SD
  vc           <- as.data.frame(VarCorr(m))
  params[i, 4] <- vc$sdcor[1]                   # SD item intercepts
  params[i, 5] <- vc$sdcor[2]                   # SD subject slopes
  params[i, 6] <- vc$sdcor[3]                   # SD subject intercepts
  params[i, 7] <- vc$sdcor[4]                   # rho (subj intercept-slope correlation)
}
```

```{r}
#| label: fig-parameter-recovery-lognormal
#| echo: false
#| eval: true
#| fig-width: 10
#| fig-height: 8
#| fig-cap: "Parameter recovery for the log-normal model with n = 20, k = 20, 1000 simulations. Red lines mark the true generating values."

op <- par(mfrow = c(3, 4), pty = "s")
hist(params[, 1], main = "b0 (Intercept)",           col = "lightblue"); abline(v = 6.59,  col = "red", lwd = 2)
hist(params[, 2], main = "b1 (Effect)",              col = "lightblue"); abline(v = 0.10,  col = "red", lwd = 2)
hist(params[, 3], main = "sigma (Residuals)",        col = "lightblue"); abline(v = 0.25,  col = "red", lwd = 2)
hist(params[, 4], main = "sigma_w (Item Intercepts)",col = "lightblue"); abline(v = 0.10,  col = "red", lwd = 2)
hist(params[, 5], main = "sigma_u1 (Subj Slopes)",   col = "lightblue"); abline(v = 0.02,  col = "red", lwd = 2)
hist(params[, 6], main = "sigma_u0 (Subj Intercepts)",col= "lightblue"); abline(v = 0.20,  col = "red", lwd = 2)
hist(params[, 7], main = "rho (Correlation)",        col = "lightblue"); abline(v = -0.50, col = "red", lwd = 2)
par(op)
```

The fixed effects and large variance components recover cleanly. The
correlation $\rho$ is still the noisiest parameter, as we saw in Part 5.

# Type-I error check

A useful sanity check on any new generative process: when the true effect is
**zero**, what proportion of simulations falsely declare significance? The
answer should be approximately $\alpha = 0.05$.

```{r}
#| label: type1-error-rate-check
#| echo: true
#| eval: true

nsim  <- 1000
pvals <- rep(NA, nsim)

for (i in 1:nsim) {
  sim_dat <- gendat_ldt_crossed_log(n = 20, k = 20,
                                    b0 = 6.59, b1 = 0,         # NO EFFECT
                                    sigma_u0 = 0.2, sigma_u1 = 0.0, # no slope variability
                                    rho = 0.0, sigma_w = 0.1, sigma = 0.25)

  m <- lmer(log(rt) ~ cond.cod + (1 + cond.cod | subj) + (1 | item), sim_dat,
            control = lmerControl(calc.derivs = FALSE))
  pvals[i] <- summary(m)$coefficients[2, 5]
}

type1_error <- mean(pvals < 0.05)
type1_error
```

If the value is close to 0.05, the model and the generative process are
consistent. Substantially higher would suggest the inference is anti-
conservative; substantially lower, the model is conservative.

# Summary and next step

In this part we:

- Recognised that real RTs are right-skewed and not well-described by a Gaussian.
- Fit `log(RT)` instead of `RT` to obtain better-behaved residuals.
- Walked through the back-transformation from log(ms) to ms and showed that the
  ms-scale effect depends on the baseline RT.
- Built `gendat_ldt_crossed_log()` to simulate from a log-normal generative
  process.
- Verified power, parameter recovery, and the Type-I error rate.

::: callout-tip
For an interactive comparison of RT distributions (log-normal, ex-Gaussian,
shifted log-normal, etc.), see Jonas K. Lindeløv's
[shiny app](https://lindeloev.shinyapps.io/shiny-rt/).
:::

The final form of the data-generating process is:

$$
\log(rt_{ij}) = \beta_0 + u_{0i} + w_{0j} + (\beta_1 + u_{1i}) \times \text{cond.cod}_{ij} + \varepsilon_{ij}
$$

The next (final) part of the course **iterates the power analysis across a grid
of sample sizes and effect sizes**, with confidence intervals on each estimate.
This is the standard deliverable for a sample-size justification in a research
proposal or pre-registration.

# Session information

```{r}
#| label: session-info
#| echo: false
#| eval: true
#| code-fold: true

sessionInfo()
```
