---
title: "Simulation-Based Power Analysis"
subtitle: "Part 8: Power analysis across varying sample sizes and effect sizes — Handout"
author:
  - name: "Filip Andras, David López-García & David Sánchez Casasola"
    affiliations:
      - name: "Centro de Investigación Mente Cerebro y Comportamiento (CIMCYC), Universidad de Granada"
date: "26 May 2026"
date-format: "D MMMM YYYY"
format:
  html:
    theme: cosmo
    toc: true
    toc-location: left
    toc-depth: 3
    embed-resources: true
    code-fold: false
    code-overflow: wrap
    code-tools: false
    fig-align: center
    number-sections: true
    page-layout: full
    include-in-header:
      - text: |
          <style>
            pre { white-space: pre-wrap; word-wrap: break-word; }
            .quarto-title-block { max-width: 1200px; margin: auto; padding: 0 2em; }
            main, #quarto-document-content { max-width: 1200px; margin: auto; padding: 0 2em; }
            h2 { margin-top: 2em; }
            h3 { margin-top: 1.5em; }
            #TOC, #quarto-margin-sidebar { width: 250px; }
          </style>
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| echo: false
#| eval: true

# load packages (suppress startup messages from lme4, lmerTest, etc.)
suppressPackageStartupMessages({
  library(tictoc)
  library(MASS)
  library(lme4)
  library(lmerTest)
  library(ggplot2)
})

options(scipen = 999) # prevents scientific notation
options(width = 120)  # wider console output for HTML rendering
set.seed(42)          # set seed for reproducibility
```

# From a point estimate to a power curve

Up to Part 7 we have computed power at **one specific combination** of design
parameters (n, effect size, number of items, random-effects variances). That is
useful but limited.

The standard deliverable for a sample-size justification — in a research
proposal, a pre-registration, or a grant — is a **power curve**: power
estimates across a range of sample sizes, often for several plausible effect
sizes.

In this part we build:

- A **first power grid**: power as a function of N subjects, for three effect
  sizes.
- A **second power grid**: power as a function of both N subjects and N items,
  for a fixed (small) effect size.

We also add **95% confidence intervals on every power estimate**, using the
Clopper-Pearson exact binomial method. Each power value is itself the mean of
1000 binary outcomes (significant / not significant), so it has its own
sampling uncertainty.

# Recap of the generative model and one-dataset workflow

The data-generating function from Part 7:

```{r}
#| label: declare-model-parameters
#| echo: true
#| eval: true

# 7 model parameters (all on log scale):
b0       <- 6.59  # grand mean log(RT); exp(6.59) ~ 730 ms
b1       <- 0.10  # log-scale frequency effect (~70 ms on RT scale)
sigma_u0 <- 0.2   # SD of subject random intercepts
sigma_u1 <- 0.02  # SD of subject random slopes
rho      <- -0.5  # correlation between subject intercepts and slopes
sigma_w  <- 0.1   # SD of item random intercepts
sigma    <- 0.25  # residual SD
```

We will walk through generating one dataset step by step to make sure the
pieces are clear, then wrap everything in `gendat_ldt_crossed_log()` and start
iterating.

```{r}
#| label: redefine-gendat-ldt-crossed-log
#| echo: true
#| eval: true

gendat_ldt_crossed_log <- function(n        = 20,
                                   k        = 20,
                                   b0       = 6.59,
                                   b1       = 0.10,
                                   sigma_u0 = 0.2,
                                   sigma_u1 = 0.02,
                                   rho      = -0.5,
                                   sigma_w  = 0.1,
                                   sigma    = 0.25) {

  sim_dat          <- expand.grid(item = 1:(k * 2), subj = 1:n)
  sim_dat$cond.cod <- rep(c(-0.5, 0.5), each = k)

  Sigma_u <- matrix(c(sigma_u0^2,                  rho * sigma_u0 * sigma_u1,
                      rho * sigma_u0 * sigma_u1,   sigma_u1^2),
                    nrow = 2)
  u <- mvrnorm(n = length(unique(sim_dat$subj)), mu = c(0, 0), Sigma = Sigma_u)
  w <- rnorm(n  = length(unique(sim_dat$item)), mean = 0, sd = sigma_w)

  N  <- dim(sim_dat)[1]
  rt <- rep(NA, N)
  for (i in 1:N) {
    rt[i] <- rlnorm(1,
                    b0 +
                    u[sim_dat[i, ]$subj, 1] +
                    w[sim_dat[i, ]$item] +
                    (b1 + u[sim_dat[i, ]$subj, 2]) * sim_dat[i, ]$cond.cod,
                    sigma)
  }

  sim_dat$rt   <- rt
  sim_dat$subj <- factor(sim_dat$subj)
  sim_dat$item <- factor(sim_dat$item)
  sim_dat
}
```

## Sanity check on the generator

```{r}
#| label: fig-sanity-hist-lognormal
#| echo: true
#| eval: true
#| fig-cap: "Simulated RTs from the log-normal generator (n = 100, k = 100 items per condition). Right-skewed as expected."

sim_dat <- gendat_ldt_crossed_log(n = 100, k = 100)
hist(sim_dat$rt, breaks = 50,
     main = "Distribution of Simulated RTs",
     xlab = "RT (ms)")
```

## Optional: a fancier visualisation

A raincloud plot combining density, boxplot, subject means, and individual
trajectories:

```{r}
#| label: fig-raincloud-plot
#| echo: false
#| eval: true
#| fig-width: 9
#| fig-height: 6
#| fig-cap: "Raincloud plot of simulated log(RT) by condition. Half-violin shows the trial-level density, boxplot the IQR, points are subject means, lines connect each subject's mean across conditions, and the red diamond + line marks the grand mean per condition."

library(ggdist)
library(dplyr)

sim_dat$condition <- ifelse(sim_dat$cond.cod == -0.5, "High Freq", "Low Freq")
sim_dat$condition <- factor(sim_dat$condition, levels = c("High Freq", "Low Freq"))

sim_dat_subj <- sim_dat |>
  group_by(subj, condition) |>
  summarize(mean_log_rt = mean(log(rt)), .groups = "drop")

set.seed(42)
sim_dat_subj <- sim_dat_subj |>
  group_by(condition) |>
  mutate(x_jitter = runif(n(), -0.01, 0.01),
         x_nudge  = if_else(condition == "High Freq", 0.30, -0.30),
         x_pos    = as.numeric(factor(condition, levels = c("High Freq", "Low Freq"))) + x_nudge + x_jitter) |>
  ungroup()

sim_dat_means <- sim_dat_subj |>
  group_by(condition) |>
  summarize(grand_mean = mean(mean_log_rt), .groups = "drop")

ggplot(sim_dat_subj, aes(x = condition, y = mean_log_rt, fill = condition)) +
  stat_halfeye(data = sim_dat |> filter(condition == "High Freq"),
               aes(x = condition, y = log(rt)),
               adjust = 0.5, width = 0.6, .width = 0, justification = 0,
               point_colour = NA, alpha = 0.5, normalize = "xy", side = "left",
               position = position_nudge(x = -0.34)) +
  stat_halfeye(data = sim_dat |> filter(condition == "Low Freq"),
               aes(x = condition, y = log(rt)),
               adjust = 0.5, width = 0.6, .width = 0, justification = 0,
               point_colour = NA, alpha = 0.5, normalize = "xy", side = "right",
               position = position_nudge(x = -0.2)) +
  geom_boxplot(data = sim_dat |> filter(condition == "High Freq"),
               aes(x = condition, y = log(rt)),
               width = 0.05, outlier.shape = NA, alpha = 0.4,
               position = position_nudge(x = 0.20)) +
  geom_boxplot(data = sim_dat |> filter(condition == "Low Freq"),
               aes(x = condition, y = log(rt)),
               width = 0.05, outlier.shape = NA, alpha = 0.4,
               position = position_nudge(x = -0.20)) +
  geom_line(aes(x = x_pos, group = subj),
            alpha = 0.3, color = "gray40", linewidth = 0.3) +
  geom_point(aes(x = x_pos, color = condition), size = 2.5, alpha = 0.7) +
  geom_line(data = sim_dat_means |>
              mutate(x_nudge = if_else(condition == "High Freq", 0.20, -0.20),
                     x_pos   = as.numeric(factor(condition, levels = c("High Freq", "Low Freq"))) + x_nudge),
            aes(x = x_pos, y = grand_mean, group = 1),
            color = "red", linewidth = 1.2, alpha = 0.9) +
  geom_point(data = sim_dat_means |>
               mutate(x_nudge = if_else(condition == "High Freq", 0.20, -0.20),
                      x_pos   = as.numeric(factor(condition, levels = c("High Freq", "Low Freq"))) + x_nudge),
             aes(x = x_pos, y = grand_mean),
             shape = 23, size = 3, fill = "red", color = "darkred", stroke = 1) +
  scale_fill_manual(values  = c("High Freq" = "steelblue", "Low Freq" = "coral")) +
  scale_color_manual(values = c("High Freq" = "steelblue", "Low Freq" = "coral")) +
  scale_y_continuous(name = "log(RT)",
                     sec.axis = sec_axis(~ exp(.), name = "RT (ms)",
                                         breaks = c(200, 400, 600, 800, 1000, 1200, 1400, 1600, 1800, 2000))) +
  labs(title = "Simulated Reaction Times by Lexical Frequency",
       subtitle = "Distributions, boxplots, subject means, and individual trajectories",
       x = "Lexical Frequency") +
  theme_minimal() +
  theme(legend.position = "none")
```

## Sanity check on the model

Fit the model once and verify the fixed-effect estimates are near the true
generating values:

```{r}
#| label: lmer-fit-sanity-check
#| echo: true
#| eval: true

m <- lmer(log(rt) ~ cond.cod + (1 + cond.cod | subj) + (1 | item), sim_dat,
          control = lmerControl(calc.derivs = FALSE))
summary(m)
```

# Two power grids

We build two complementary power grids, presented in parallel tabs below.

The **first grid** varies subjects × effect sizes (with items fixed) — this is
the standard "how many subjects do I need at each plausible effect size?"
question. The **second grid** varies subjects × items (with effect size fixed
at the small value) — useful when stimulus pool size is a design lever you
control, common in psycholinguistics.

Both grids share the same machinery: build a parameter grid, run a nested
power-simulation loop, load the precomputed result, compute Clopper-Pearson
confidence intervals on each power estimate, and visualise. Each tab is
self-contained.

::: callout-note
**Confidence intervals on power.** Each power estimate is the proportion of
`nsim` independent significant results — a binomial proportion. The
Clopper-Pearson method gives an *exact* confidence interval for that
proportion. We define the helper `calculate_power_ci()` once here and reuse it
in both tabs.
:::

```{r}
#| label: define-calculate-power-ci
#| echo: true
#| eval: true
#| code-fold: true
#| code-summary: "Show code (CI helper function)"

calculate_power_ci <- function(n_significant, n_total, conf.level = 0.95) {
  if (is.na(n_significant) || is.na(n_total) || n_total == 0) {
    return(c(power = NA, lower = NA, upper = NA))
  }
  n_significant <- as.integer(round(n_significant))
  n_total       <- as.integer(round(n_total))
  if (n_significant < 0)        n_significant <- 0
  if (n_significant > n_total)  n_significant <- n_total

  test_result <- binom.test(n_significant, n_total, conf.level = conf.level)
  c(power = unname(test_result$estimate),
    lower = test_result$conf.int[1],
    upper = test_result$conf.int[2])
}
```

::: panel-tabset

## First grid: subjects × effect sizes

The first power curve varies the number of subjects across three effect sizes,
with items fixed at 40 per condition:

```{r}
#| label: build-power-grid-1d
#| echo: true
#| eval: true

n_subj       <- c(24, 36, 48, 60, 72, 84, 96)
effect_sizes <- c(0.06, 0.10, 0.14)
n_item       <- 40  # fixed number of items per condition

power_grid       <- expand.grid(n = n_subj, b1 = effect_sizes)
power_grid$power <- NA
power_grid
```

Each row of `power_grid` is one (n, effect) combination. We will fill in the
`power` column by running `nsim = 1000` simulations per row.

```{r}
#| label: total-simulations-budget
#| echo: true
#| eval: true

nsim <- 1000
nrow(power_grid)         # number of (n, effect) combinations
nrow(power_grid) * nsim  # total number of model fits
```

::: callout-warning
21,000 model fits is a lot of computation. The full first grid takes
**~5.34 hours** on a standard laptop. Below we show the loop code (set to
`eval: false`) and instead load the precomputed result.
:::

### The simulation loop

```{r}
#| label: power-grid-1d-nested-loops
#| echo: true
#| eval: false

tic()
for (row in 1:nrow(power_grid)) {
  n_subj <- power_grid$n[row]
  effect <- power_grid$b1[row]

  message(sprintf("Running: n=%d, b1=%.2f (%d/%d)",
                  n_subj, effect, row, nrow(power_grid)))

  pvals <- rep(NA, nsim)
  for (i in 1:nsim) {
    sim_dat <- gendat_ldt_crossed_log(n = n_subj, k = n_item,
                                      b0 = 6.59, b1 = effect,
                                      sigma_u0 = 0.2, sigma_u1 = 0.02,
                                      rho = -0.5, sigma_w = 0.1, sigma = 0.25)
    m <- lmer(log(rt) ~ cond.cod + (1 + cond.cod | subj) + (1 | item), sim_dat,
              control = lmerControl(calc.derivs = FALSE))
    pvals[i] <- summary(m)$coefficients[2, 5]
  }
  power_grid$power[row] <- mean(pvals < 0.05)
}
toc() # 19246.916 sec elapsed = 5.34h

# save(power_grid, file = "../data/processed_data/power_grid_sub_eff.RData")
```

### Load the precomputed result

```{r}
#| label: load-power-grid-1d
#| echo: true
#| eval: true

load("../data/processed_data/power_grid_sub_eff.RData")
power_grid
```

### Add confidence intervals

```{r}
#| label: compute-ci-power-grid-1d
#| echo: true
#| eval: true
#| code-fold: true
#| code-summary: "Show code (CI computation)"

power_grid$n_sig    <- round(power_grid$power * nsim)
power_grid$lower_ci <- NA
power_grid$upper_ci <- NA

for (row in 1:nrow(power_grid)) {
  ci <- calculate_power_ci(power_grid$n_sig[row], nsim)
  power_grid$lower_ci[row] <- ci["lower"]
  power_grid$upper_ci[row] <- ci["upper"]
}

print(power_grid[, c("n", "b1", "power", "lower_ci", "upper_ci")])
```

The CIs are narrow because `nsim = 1000` is large. More simulations would give
narrower CIs but at the cost of more compute time.

### Visualisation: power curves

```{r}
#| label: fig-power-curves
#| echo: true
#| eval: true
#| code-fold: true
#| code-summary: "Show code (ggplot recipe)"
#| fig-width: 10
#| fig-height: 7
#| fig-cap: "Power curves with 95% Clopper-Pearson CIs for three effect sizes, as a function of subject sample size (40 items per condition)."

power_grid$effect_label <- factor(power_grid$b1,
                                  levels = c(0.06, 0.10, 0.14),
                                  labels = c("Small (b1=0.06)",
                                             "Medium (b1=0.10)",
                                             "Large (b1=0.14)"))

ggplot(power_grid, aes(x = n, y = power,
                       color = effect_label, fill = effect_label,
                       group = effect_label)) +
  geom_ribbon(aes(ymin = lower_ci, ymax = upper_ci),
              alpha = 0.2, color = NA) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  geom_hline(yintercept = 0.80, linetype = "dashed", color = "black", linewidth = 0.8) +
  geom_hline(yintercept = 0.95, linetype = "dashed", color = "black", linewidth = 0.8) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.1)) +
  scale_x_continuous(breaks = n_subj) +
  scale_color_manual(values = c("#E69F00", "#56B4E9", "#009E73")) +
  scale_fill_manual(values  = c("#E69F00", "#56B4E9", "#009E73")) +
  labs(title    = "Power Curves with 95% Confidence Intervals",
       subtitle = sprintf("%d simulations per point", nsim),
       x        = "Number of Subjects",
       y        = "Statistical Power (1 - β)",
       color    = "Effect Size",
       fill     = "Effect Size") +
  theme_minimal(base_size = 14) +
  theme(legend.position  = "bottom",
        plot.title       = element_text(face = "bold"),
        panel.grid.minor = element_blank())
```

How to read this plot:

- **x-axis**: number of subjects.
- **y-axis**: power. Dashed lines mark the 80% and 95% conventional targets.
- **Three curves**: power as a function of subjects for the three effect sizes.
- **Shaded ribbons**: 95% Clopper-Pearson CIs on each power estimate.

A larger effect needs fewer subjects. Smaller effects need many more. The
curve shape (S-curve flattening out near 1) is typical.

## Second grid: subjects × items (small effect)

What if we instead want to understand how the **number of items per condition**
trades off against the number of subjects, for a fixed (small) effect?

```{r}
#| label: build-power-grid-2d
#| echo: true
#| eval: true

sample_sizes_2d <- c(24, 36, 48, 60, 72, 84, 96)
item_numbers_2d <- c(40, 60, 80, 100, 120)
effect_fixed    <- 0.06

power_grid_2d       <- expand.grid(n = sample_sizes_2d, nitem = item_numbers_2d)
power_grid_2d$power <- NA

nsim <- 500 # fewer simulations per cell for this larger grid
nrow(power_grid_2d) * nsim  # total simulations
```

::: callout-warning
This second grid covers $7 \times 5 = 35$ cells with 500 simulations each — 17,500
model fits. It took **~8.37 hours** to run, so once again we show the loop and
load the precomputed result.
:::

### The simulation loop

```{r}
#| label: power-grid-2d-nested-loops
#| echo: true
#| eval: false

tic()
for (row in 1:nrow(power_grid_2d)) {
  n_subj  <- power_grid_2d$n[row]
  n_items <- power_grid_2d$nitem[row]

  message(sprintf("Running: n=%d, k=%d (%d/%d)",
                  n_subj, n_items, row, nrow(power_grid_2d)))

  pvals <- rep(NA, nsim)
  for (i in 1:nsim) {
    sim_dat <- gendat_ldt_crossed_log(n = n_subj, k = n_items,
                                      b0 = 6.59, b1 = effect_fixed,
                                      sigma_u0 = 0.2, sigma_u1 = 0.02,
                                      rho = -0.5, sigma_w = 0.1, sigma = 0.25)
    m <- lmer(log(rt) ~ cond.cod + (1 + cond.cod | subj) + (1 | item), sim_dat,
              control = lmerControl(calc.derivs = FALSE))
    pvals[i] <- summary(m)$coefficients[2, 5]
  }
  power_grid_2d$power[row] <- mean(pvals < 0.05)
}
toc() # 30145.484 sec elapsed = 8.37h

# save(power_grid_2d, file = "../data/processed_data/power_grid_2d.RData")
```

### Load the precomputed result

```{r}
#| label: load-power-grid-2d
#| echo: true
#| eval: true

load("../data/processed_data/power_grid_2d.RData")
power_grid_2d
```

### Add confidence intervals

```{r}
#| label: compute-ci-power-grid-2d
#| echo: true
#| eval: true
#| code-fold: true
#| code-summary: "Show code (CI computation)"

nsim <- 500 # matches the simulation budget for the second grid

power_grid_2d$n_sig    <- round(power_grid_2d$power * nsim)
power_grid_2d$lower_ci <- NA
power_grid_2d$upper_ci <- NA

for (row in 1:nrow(power_grid_2d)) {
  if (!is.na(power_grid_2d$power[row])) {
    ci <- calculate_power_ci(power_grid_2d$n_sig[row], nsim)
    power_grid_2d$lower_ci[row] <- ci["lower"]
    power_grid_2d$upper_ci[row] <- ci["upper"]
  }
}

power_grid_2d[, c("n", "nitem", "power", "lower_ci", "upper_ci")]
```

### Visualisation: heatmap

```{r}
#| label: fig-power-heatmap
#| echo: true
#| eval: true
#| code-fold: true
#| code-summary: "Show code (ggplot recipe)"
#| fig-width: 10
#| fig-height: 7
#| fig-cap: "Heatmap of power as a function of subjects and items, for a small effect (b1 = 0.06 on the log scale). Green cells are at or above 80% power; orange/red cells are below. Bracket values are the 95% Clopper-Pearson CIs."

ggplot(power_grid_2d, aes(x = factor(n), y = factor(nitem), fill = power)) +
  geom_tile(color = "white", size = 1) +
  geom_text(aes(label = sprintf("%.2f", power)),
            color = "black", size = 5, fontface = "bold", vjust = -0.5) +
  geom_text(aes(label = sprintf("[%.2f, %.2f]", lower_ci, upper_ci)),
            color = "grey40", size = 3, vjust = 1.2) +
  scale_fill_gradient2(low      = "#d73027",
                       mid      = "#fee08b",
                       high     = "#1a9850",
                       midpoint = 0.80,
                       limits   = c(0, 1),
                       breaks   = seq(0, 1, 0.2),
                       name     = "Power") +
  labs(title    = "Power as a Function of Subjects and Items",
       subtitle = sprintf("Effect size: b1 = %.2f (log scale), %d simulations per point",
                          effect_fixed, nsim),
       x = "Number of Subjects",
       y = "Number of Items") +
  theme_minimal(base_size = 14) +
  theme(plot.title  = element_text(face = "bold"),
        panel.grid  = element_blank())
```

The heatmap shows that **both** axes matter. To detect a small effect with
adequate power, you can either run many subjects with moderate items, or fewer
subjects with many items. The corner of the grid (96 subjects × 120 items per
condition) is the easiest case.

:::

# What to do next

The power analysis above is intentionally coarse — three effect sizes in the
first grid, seven subject counts × five item counts in the second. For an
actual study planning
exercise:

- **Zoom in** on the range of (subjects, items) where you cross the 80% or
  95% threshold.
- **Run more simulations per cell** (e.g., 2000+ instead of 500) to tighten the
  CIs.
- **Refine the input parameters** ($\sigma_{u0}, \sigma_{u1}, \rho, \sigma_w,
  \sigma$) from a pilot dataset rather than relying on values from the
  literature — they are the single biggest source of uncertainty in the result.

# The final model

After eight parts we have built up a full data-generating process and a
matching LMM:

$$
\log(rt_{ij}) = \beta_0 + u_{0i} + w_{0j} + (\beta_1 + u_{1i}) \times \text{cond.cod}_{ij} + \varepsilon_{ij}
$$

where $\varepsilon_{ij} \sim \text{Normal}(0, \sigma)$,

$$
\begin{pmatrix} u_0 \\ u_1 \end{pmatrix} \sim \text{MVN} \left(
\begin{pmatrix} 0 \\ 0 \end{pmatrix}, \, \Sigma_u \right), \quad
\Sigma_u = \begin{pmatrix}
\sigma_{u0}^2 & \rho \, \sigma_{u0} \sigma_{u1} \\
\rho \, \sigma_{u0} \sigma_{u1} & \sigma_{u1}^2
\end{pmatrix},
$$

and $w_0 \sim \text{Normal}(0, \sigma_w)$.

Corresponding `lmer()` formula:

```r
lmer(log(rt) ~ cond.cod + (1 + cond.cod | subj) + (1 | item), data = ...)
```

# Where to go next

This course covered simulation-based power analysis in R using `lme4`. Two
recommended next steps:

- **Vasishth & Nicenboim**, *Using simulation to understand your model* —
  [Freq_CogSci](https://vasishth.github.io/Freq_CogSci/using-simulation-to-understand-your-model.html).
  Same paradigm, more examples, written for psycholinguistics.
- **Julia + MixedModelsSim.jl** — [simulation
  tutorial](https://repsychling.github.io/MixedModelsSim.jl/stable/simulation_tutorial/).
  Julia is roughly 10-100× faster than R for these simulations, which makes it
  feasible to scan much larger grids.

# Session information

```{r}
#| label: session-info
#| echo: false
#| eval: true
#| code-fold: true

sessionInfo()
```
