---
title: "Simulation-Based Power Analysis"
subtitle: "Part 9: Recap — generative formulas and gendat functions across Parts 1-8 — Handout"
author:
  - name: "Filip Andras, David López-García & David Sánchez Casasola"
    affiliations:
      - name: "Centro de Investigación Mente Cerebro y Comportamiento (CIMCYC), Universidad de Granada"
date: "26 May 2026"
date-format: "D MMMM YYYY"
format:
  html:
    theme: cosmo
    toc: true
    toc-location: left
    toc-depth: 3
    embed-resources: true
    code-fold: false
    code-overflow: wrap
    code-tools: false
    fig-align: center
    number-sections: true
    page-layout: full
    include-in-header:
      - text: |
          <style>
            pre { white-space: pre-wrap; word-wrap: break-word; }
            .quarto-title-block { max-width: 1200px; margin: auto; padding: 0 2em; }
            main, #quarto-document-content { max-width: 1200px; margin: auto; padding: 0 2em; }
            h2 { margin-top: 2em; }
            h3 { margin-top: 1.5em; }
            #TOC, #quarto-margin-sidebar { width: 250px; }
          </style>
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| echo: false
#| eval: true

suppressPackageStartupMessages({
  library(tictoc)
  library(MASS)
  library(lme4)
  library(lmerTest)
  library(ggplot2)
})

options(scipen = 999)
options(width = 120)
set.seed(42)
```

# What we built across Parts 1-8

Each part of the course added **one new piece** to the data-generating process,
and the corresponding R function (`gendat_*()`) grew alongside the formula.

This recap lays the formulas and the generator functions side by side using
tabsets. Pick a part in either tabset to see what was new at that step.

# Overview

The table below maps each part to the new modelling ingredient it introduced
and to the function that implements it:

| Part | New ingredient | Function |
|------|----------------|----------|
| 1    | Between-subjects design, t-test | `betwsubj_simdat_ldt()` |
| 1.2  | Linear-model parametrisation (same data) | `betwsubj_simdat_ldt()` |
| 2    | Within-subjects + random intercept ($u_{0i}$) | `withinsubj_simdat()` |
| 3    | Multiple trials per cell | (inline, no wrapper) |
| 4    | Random slope ($u_{1i}$) | `gendat_ldt()` |
| 5    | Correlated $u_{0i}, u_{1i}$ via $\rho$ | `gendat_ldt_corr()` |
| 6    | Item random intercept ($w_{0j}$), fully crossed design | `gendat_ldt_crossed()` |
| 7    | Log-normal RT | `gendat_ldt_crossed_log()` |
| 8    | Iterate the Part 7 generator over a grid | (same generator) |

The next two tabsets show the generative formula and the function side by side
for each part.

# Generative formulas

::: panel-tabset

## Part 1

**Between-subjects, independent observations:**

$$rt_i = \mu_c + \varepsilon_i$$

where $\mu_c$ is the condition mean (700 or 750 ms) and
$\varepsilon_i \sim N(0, 200)$. There is no random structure across observations
— each $rt$ depends only on which group the subject was assigned to.

## Part 1.2

**Same data, linear-model parametrisation (effects coding):**

$$rt_i = \beta_0 + \beta_1 \cdot \text{cond.cod}_i + \varepsilon_i$$

with $\beta_0 = 725$ (grand mean), $\beta_1 = 50$ (full difference between
conditions), and $\varepsilon_i \sim N(0, 200)$.

Nothing changes in the data — only how we parametrise it for `lm()`.

## Part 2

**Within-subjects + random intercept:**

$$rt_{ij} = \beta_0 + u_{0i} + \beta_1 \cdot \text{cond.cod}_{ij} + \varepsilon_{ij}$$

with $u_{0i} \sim N(0, 150)$. Each subject now has their own intercept, drawn
once and used for both observations from that subject. This creates the
within-subjects dependency.

## Part 3

**Multiple trials per cell (same formula, more rows):**

$$rt_{ijk} = \beta_0 + u_{0i} + \beta_1 \cdot \text{cond.cod}_{ij} + \varepsilon_{ijk}$$

$k$ trials per (subject, condition) cell. The formula is unchanged; the index
$k$ just enumerates trials. The nested `rep()` calls in the code align each
subject's intercept with their $2k$ trials.

## Part 4

**+ random slope:**

$$rt_{ij} = \beta_0 + u_{0i} + (\beta_1 + u_{1i}) \cdot \text{cond.cod}_{ij} + \varepsilon_{ij}$$

with $u_{1i} \sim N(0, 20)$, drawn **independently** of $u_{0i}$. Each subject
now also has their own slope — the size of the frequency effect varies across
subjects.

## Part 5

**Correlated random intercepts and slopes:**

$$rt_{ij} = \beta_0 + u_{0i} + (\beta_1 + u_{1i}) \cdot \text{cond.cod}_{ij} + \varepsilon_{ij}$$

$$\begin{pmatrix} u_0 \\ u_1 \end{pmatrix} \sim \text{MVN}\left( \mathbf{0}, \, \Sigma_u \right), \quad \Sigma_u = \begin{pmatrix} \sigma_{u0}^2 & \rho \, \sigma_{u0} \sigma_{u1} \\ \rho \, \sigma_{u0} \sigma_{u1} & \sigma_{u1}^2 \end{pmatrix}$$

The two random effects are now drawn from a **bivariate normal** controlled by
the correlation $\rho$. With $\rho < 0$, slower subjects tend to show smaller
effects.

## Part 6

**+ fully crossed design with item random intercept:**

$$rt_{ij} = \beta_0 + u_{0i} + w_{0j} + (\beta_1 + u_{1i}) \cdot \text{cond.cod}_{ij} + \varepsilon_{ij}$$

with $w_{0j} \sim N(0, \sigma_w)$. The item index $j$ now varies independently
of the subject index $i$ (fully crossed) and each item has its own random
intercept.

## Part 7

**Log-normal RT (positive, right-skewed):**

$$\log(rt_{ij}) = \beta_0 + u_{0i} + w_{0j} + (\beta_1 + u_{1i}) \cdot \text{cond.cod}_{ij} + \varepsilon_{ij}$$

All parameters now live on the **log scale** (so $b_0 \approx 6.59$ corresponds
to $\exp(6.59) \approx 730$ ms). Back-transform individual predictions with
`exp()`.

## Part 8

**Same generator as Part 7, iterated over a grid:**

For each $(n, b_1)$ combination (first grid) or each $(n, k)$ combination
(second grid):

1. Generate `nsim` independent datasets with the Part 7 process.
2. Fit the LMM and extract the $p$-value for the condition effect.
3. Compute power as the proportion of $p < 0.05$.
4. Wrap the result with a Clopper-Pearson 95% confidence interval.

:::

# Generator functions

::: panel-tabset

## Part 1 / 1.2

```{r}
#| label: gendat-part1
#| echo: true
#| eval: false

betwsubj_simdat_ldt <- function(n = 10,
                                mean_freq_high = 700,
                                mean_freq_low  = 750,
                                stdev     = 200) {
  freq_high <- rnorm(n, mean = mean_freq_high, sd = stdev)
  freq_low  <- rnorm(n, mean = mean_freq_low,  sd = stdev)
  subj      <- 1:(2 * n)
  cond      <- factor(rep(c("freq_high", "freq_low"), each = n))
  cond.cod  <- ifelse(cond == "freq_high", -0.5, 0.5)
  data.frame(subj = subj, cond = cond, cond.cod = cond.cod,
             rt = c(freq_high, freq_low))
}
```

Two independent `rnorm()` calls, one per condition. No within-subject
dependency.

## Part 2

```{r}
#| label: gendat-part2
#| echo: true
#| eval: false

withinsubj_simdat <- function(n = 10,
                              b0 = 725,
                              b1 = 50,
                              sigma_u0 = 150,
                              sigma = 200) {
  u0       <- round(rnorm(n, mean = 0, sd = sigma_u0))
  cond.cod <- rep(c(-0.5, 0.5), each = n)
  subj     <- rep(1:n, times = 2)
  rt <- round(b0 +
              rep(u0, times = 2) +              # subject adjustment in BOTH conds
              b1 * cond.cod +
              rnorm(2 * n, mean = 0, sd = sigma))
  data.frame(subj = subj, cond.cod = cond.cod, rt = rt)
}
```

The `rep(u0, times = 2)` trick is what creates the within-subjects dependency.

## Part 3

**No wrapper — extends Part 2 inline by repeating each subject `k` times:**

```{r}
#| label: gendat-part3
#| echo: true
#| eval: false

n <- 20  # subjects
k <- 20  # trials per condition

cond.cod <- rep(c(-0.5, 0.5), each = n * k)
subj     <- rep(rep(1:n, each = k), times = 2)
sim_dat  <- data.frame(subj = subj, cond.cod = cond.cod)

sim_dat$rt <- 725 +
              rep(rep(round(rnorm(n, 0, 150)), each = k), 2) + # subject random intercept
              50 * cond.cod +                                  # condition effect
              rnorm(n * k * 2, 0, 200)                         # residual
```

The nested `rep(rep(..., each = k), 2)` aligns each subject's intercept across
their $2k$ trials.

## Part 4

```{r}
#| label: gendat-part4
#| echo: true
#| eval: false

gendat_ldt <- function(n = 20, k = 20,
                       b0 = 725, b1 = 50,
                       sigma_u0 = 150, sigma_u1 = 20,
                       sigma = 200) {
  u <- data.frame(subject = 1:n,
                  u0 = rnorm(n, 0, sigma_u0),
                  u1 = rnorm(n, 0, sigma_u1))
  cond.cod <- rep(c(-0.5, 0.5), each = n * k)
  subj     <- rep(rep(1:n, each = k), times = 2)
  sim_dat  <- data.frame(subj = subj, cond.cod = cond.cod)
  nrows <- dim(sim_dat)[1]
  rt    <- rep(NA, nrows)
  for (i in 1:nrows) {
    rt[i] <- b0 +
             u[sim_dat[i, ]$subj, ]$u0 +
             (b1 + u[sim_dat[i, ]$subj, ]$u1) * sim_dat[i, ]$cond.cod +
             rnorm(1, 0, sigma)
  }
  sim_dat$rt <- rt
  sim_dat
}
```

The `u` data frame holds one row per subject. The row-by-row loop looks up
each subject's adjustments via `u[sim_dat[i, ]$subj, ]`.

## Part 5

```{r}
#| label: gendat-part5
#| echo: true
#| eval: false

gendat_ldt_corr <- function(n = 20, k = 20,
                            b0 = 725, b1 = 50,
                            sigma_u0 = 150, sigma_u1 = 20,
                            rho = -0.5,                          # NEW
                            sigma = 200) {
  Sigma <- matrix(c(sigma_u0^2,                  sigma_u0 * sigma_u1 * rho,
                    sigma_u0 * sigma_u1 * rho,   sigma_u1^2), ncol = 2)
  u <- as.data.frame(mvrnorm(n = n, mu = c(0, 0), Sigma = Sigma))
  colnames(u) <- c("u0", "u1"); u$subject <- 1:n
  cond.cod <- rep(c(-0.5, 0.5), each = n * k)
  subj     <- rep(rep(1:n, each = k), times = 2)
  sim_dat  <- data.frame(subj = subj, cond.cod = cond.cod)
  nrows <- dim(sim_dat)[1]; rt <- rep(NA, nrows)
  for (i in 1:nrows) {
    rt[i] <- b0 + u[sim_dat[i, ]$subj, ]$u0 +
             (b1 + u[sim_dat[i, ]$subj, ]$u1) * sim_dat[i, ]$cond.cod +
             rnorm(1, 0, sigma)
  }
  sim_dat$rt <- rt; sim_dat
}
```

The two independent `rnorm()` calls from Part 4 collapse into one
`mvrnorm()` call controlled by the variance-covariance matrix.

## Part 6

```{r}
#| label: gendat-part6
#| echo: true
#| eval: false

gendat_ldt_crossed <- function(n = 20, k = 20,
                               b0 = 725, b1 = 50,
                               sigma_u0 = 150, sigma_u1 = 20, rho = -0.5,
                               sigma_w = 100,                       # NEW: item SD
                               sigma = 200) {
  sim_dat          <- expand.grid(item = 1:(k*2), subj = 1:n)        # fully crossed
  sim_dat$cond.cod <- rep(c(-0.5, 0.5), each = k)
  Sigma_u <- matrix(c(sigma_u0^2,                  rho * sigma_u0 * sigma_u1,
                      rho * sigma_u0 * sigma_u1,   sigma_u1^2), nrow = 2)
  u <- mvrnorm(n = length(unique(sim_dat$subj)), mu = c(0, 0), Sigma = Sigma_u)
  w <- rnorm(n = length(unique(sim_dat$item)), mean = 0, sd = sigma_w) # NEW: item RE
  N <- dim(sim_dat)[1]; rt <- rep(NA, N)
  for (i in 1:N) {
    rt[i] <- rnorm(1,
                   b0 + u[sim_dat[i, ]$subj, 1] + w[sim_dat[i, ]$item] +
                   (b1 + u[sim_dat[i, ]$subj, 2]) * sim_dat[i, ]$cond.cod,
                   sigma)
  }
  sim_dat$rt <- rt
  sim_dat$subj <- factor(sim_dat$subj); sim_dat$item <- factor(sim_dat$item)
  sim_dat
}
```

`expand.grid()` builds the fully crossed design (every subject sees every
item). The new `w` vector adds one random intercept per item.

## Part 7

```{r}
#| label: gendat-part7
#| echo: true
#| eval: false

gendat_ldt_crossed_log <- function(n = 20, k = 20,
                                   b0 = 6.59,                # log scale
                                   b1 = 0.10,                # log scale
                                   sigma_u0 = 0.2, sigma_u1 = 0.02, rho = -0.5,
                                   sigma_w = 0.1,
                                   sigma = 0.25) {           # all on log scale
  sim_dat          <- expand.grid(item = 1:(k*2), subj = 1:n)
  sim_dat$cond.cod <- rep(c(-0.5, 0.5), each = k)
  Sigma_u <- matrix(c(sigma_u0^2,                  rho * sigma_u0 * sigma_u1,
                      rho * sigma_u0 * sigma_u1,   sigma_u1^2), nrow = 2)
  u <- mvrnorm(n = length(unique(sim_dat$subj)), mu = c(0, 0), Sigma = Sigma_u)
  w <- rnorm(n = length(unique(sim_dat$item)), mean = 0, sd = sigma_w)
  N <- dim(sim_dat)[1]; rt <- rep(NA, N)
  for (i in 1:N) {
    rt[i] <- rlnorm(1,                                       # rlnorm() not rnorm()
                    b0 + u[sim_dat[i, ]$subj, 1] + w[sim_dat[i, ]$item] +
                    (b1 + u[sim_dat[i, ]$subj, 2]) * sim_dat[i, ]$cond.cod,
                    sigma)
  }
  sim_dat$rt <- rt
  sim_dat$subj <- factor(sim_dat$subj); sim_dat$item <- factor(sim_dat$item)
  sim_dat
}
```

Only one structural change from Part 6: `rnorm()` becomes `rlnorm()`. All the
parameters are interpreted on the **log scale** instead of the millisecond
scale.

## Part 8

**Same generator as Part 7. The new code is the nested grid loop that wraps
the generator:**

```{r}
#| label: gendat-part8
#| echo: true
#| eval: false

# first grid: subjects x effect size
for (row in 1:nrow(power_grid)) {
  n_subj <- power_grid$n[row]
  effect <- power_grid$b1[row]
  pvals  <- rep(NA, nsim)

  for (i in 1:nsim) {
    sim_dat <- gendat_ldt_crossed_log(n = n_subj, k = n_item, b1 = effect, ...)
    m <- lmer(log(rt) ~ cond.cod + (1 + cond.cod | subj) + (1 | item), sim_dat,
              control = lmerControl(calc.derivs = FALSE))
    pvals[i] <- summary(m)$coefficients[2, 5]
  }

  power_grid$power[row] <- mean(pvals < 0.05)
}
```

Two nested loops: the outer iterates over rows of the grid, the inner runs
`nsim` simulations and records $p$-values.

:::

# Summary

Three things to notice as you flick through the tabs:

- **Additive structure.** Every formula contains all the previous ones inside.
  Each tab adds exactly one new ingredient (a new random effect, a new
  correlation, a new transformation).
- **Function names track complexity.** From `betwsubj_simdat_ldt()` (Part 1) to
  `gendat_ldt_crossed_log()` (Part 7), the name accumulates suffixes
  (`_corr`, `_crossed`, `_log`) as each new piece is added.
- **The `lmer()` formula grows in lock-step.** Starting from
  `lm(rt ~ cond.cod)` and ending at
  `lmer(log(rt) ~ cond.cod + (1 + cond.cod | subj) + (1 | item))`. The
  match between the generative process and the analysis model is what makes
  the power estimates honest.

# Session information

```{r}
#| label: session-info
#| echo: false
#| eval: true
#| code-fold: true

sessionInfo()
```
