---
title: "Simulation-Based Power Analysis"
subtitle: "Part 9: Recap — generative formulas and gendat functions across Parts 1-8"
author:
  - name: "Filip Andras, David López-García & David Sánchez Casasola"
    affiliations:
      - name: "Centro de Investigación Mente Cerebro y Comportamiento (CIMCYC), Universidad de Granada"
date: "26 May 2026"
date-format: "D MMMM YYYY"
format:
  revealjs:
    theme: serif
    css: ../custom.css
    slide-number: true
    chalkboard: true
    code-fold: false
    code-overflow: wrap
    scrollable: true
    footer: "Justificación del tamaño muestral y análisis de potencia estadística | v2026.2"
execute:
  warning: false
  message: false
---

## Part 9: Recap — what we built across Parts 1-8

```{r}
#| label: setup
#| echo: false
#| eval: true

suppressPackageStartupMessages({
  library(tictoc)
  library(MASS)
  library(lme4)
  library(lmerTest)
  library(ggplot2)
})

options(scipen = 999)
options(width = 120)
set.seed(42)
```

::: incremental
- Each part added **one new piece** to the data-generating process.
- The R function (`gendat_*()`) grew alongside the formula.
- This recap lays them side by side via tabsets — pick a part, see the formula and the code.
:::

## Overview

| Part | New ingredient | Function |
|------|----------------|----------|
| 1    | Between-subjects design, t-test | `betwsubj_simdat_ldt()` |
| 1.2  | Linear model (same data) | `betwsubj_simdat_ldt()` |
| 2    | Within-subjects + random intercept ($u_{0i}$) | `withinsubj_simdat()` |
| 3    | Multiple trials per cell | (inline) |
| 4    | Random slope ($u_{1i}$) | `gendat_ldt()` |
| 5    | Correlated $u_{0i}, u_{1i}$ via $\rho$ | `gendat_ldt_corr()` |
| 6    | Item random intercept ($w_{0j}$) | `gendat_ldt_crossed()` |
| 7    | Log-normal RT | `gendat_ldt_crossed_log()` |
| 8    | Iterate over power grids | (same generator) |

## Generative formulas {background-color="#43464B"}

## Generative formulas {.smaller}

::: panel-tabset

### Part 1

**Between-subjects, independent observations:**

$$rt_i = \mu_c + \varepsilon_i$$

with $\mu_c$ = condition mean (700 or 750 ms) and $\varepsilon_i \sim N(0, 200)$.

### Part 1.2

**Same data, linear-model parametrisation (effects coding):**

$$rt_i = \beta_0 + \beta_1 \cdot \text{cond.cod}_i + \varepsilon_i$$

with $\beta_0 = 725$, $\beta_1 = 50$, $\varepsilon_i \sim N(0, 200)$.

### Part 2

**Within-subjects + random intercept:**

$$rt_{ij} = \beta_0 + u_{0i} + \beta_1 \cdot \text{cond.cod}_{ij} + \varepsilon_{ij}$$

with $u_{0i} \sim N(0, 150)$.

### Part 3

**Multiple trials per cell (same formula, more rows):**

$$rt_{ijk} = \beta_0 + u_{0i} + \beta_1 \cdot \text{cond.cod}_{ij} + \varepsilon_{ijk}$$

$k$ trials per (subject, condition).

### Part 4

**+ random slope:**

$$rt_{ij} = \beta_0 + u_{0i} + (\beta_1 + u_{1i}) \cdot \text{cond.cod}_{ij} + \varepsilon_{ij}$$

with $u_{1i} \sim N(0, 20)$, independent of $u_{0i}$.

### Part 5

**Correlated random intercepts and slopes:**

$$rt_{ij} = \beta_0 + u_{0i} + (\beta_1 + u_{1i}) \cdot \text{cond.cod}_{ij} + \varepsilon_{ij}$$

$$\begin{pmatrix} u_0 \\ u_1 \end{pmatrix} \sim \text{MVN}\left( \mathbf{0}, \, \Sigma_u \right), \quad \Sigma_u = \begin{pmatrix} \sigma_{u0}^2 & \rho \, \sigma_{u0} \sigma_{u1} \\ \rho \, \sigma_{u0} \sigma_{u1} & \sigma_{u1}^2 \end{pmatrix}$$

### Part 6

**+ fully crossed design with item random intercept:**

$$rt_{ij} = \beta_0 + u_{0i} + w_{0j} + (\beta_1 + u_{1i}) \cdot \text{cond.cod}_{ij} + \varepsilon_{ij}$$

with $w_{0j} \sim N(0, \sigma_w)$. Item index $j$ now varies independently of subject index $i$.

### Part 7

**Log-normal RT (positive, right-skewed):**

$$\log(rt_{ij}) = \beta_0 + u_{0i} + w_{0j} + (\beta_1 + u_{1i}) \cdot \text{cond.cod}_{ij} + \varepsilon_{ij}$$

All parameters now on the **log scale**. Back-transform with `exp()`.

### Part 8

**Same generator as Part 7, iterated over a grid:**

For each $(n, b_1)$ (first grid) or $(n, k)$ (second grid) combination:

- Generate `nsim` datasets with the Part 7 process.
- Fit the LMM and extract the $p$-value.
- Power = proportion of $p < 0.05$.

:::

## Generator functions {background-color="#43464B"}

## Generator functions {.smaller}

::: panel-tabset

### Part 1 / 1.2

```r
betwsubj_simdat_ldt <- function(n = 10,
                                mean_freq_high = 700,
                                mean_freq_low  = 750,
                                stdev     = 200) {
  freq_high <- rnorm(n, mean = mean_freq_high, sd = stdev)
  freq_low  <- rnorm(n, mean = mean_freq_low,  sd = stdev)
  subj      <- 1:(2 * n)
  cond      <- factor(rep(c("freq_high", "freq_low"), each = n))
  cond.cod  <- ifelse(cond == "freq_high", -0.5, 0.5)
  data.frame(subj = subj, cond = cond, cond.cod = cond.cod,
             rt = c(freq_high, freq_low))
}
```

### Part 2

```r
withinsubj_simdat <- function(n = 10,
                              b0 = 725,
                              b1 = 50,
                              sigma_u0 = 150,
                              sigma = 200) {
  u0       <- round(rnorm(n, mean = 0, sd = sigma_u0))
  cond.cod <- rep(c(-0.5, 0.5), each = n)
  subj     <- rep(1:n, times = 2)
  rt <- round(b0 +
              rep(u0, times = 2) +
              b1 * cond.cod +
              rnorm(2 * n, mean = 0, sd = sigma))
  data.frame(subj = subj, cond.cod = cond.cod, rt = rt)
}
```

### Part 3

**Inline (no wrapper) — extends Part 2 by repeating each subject `k` times:**

```r
n <- 20  # subjects
k <- 20  # trials per condition

cond.cod <- rep(c(-0.5, 0.5), each = n * k)
subj     <- rep(rep(1:n, each = k), times = 2)
sim_dat  <- data.frame(subj = subj, cond.cod = cond.cod)

sim_dat$rt <- 725 +
              rep(rep(round(rnorm(n, 0, 150)), each = k), 2) +
              50 * cond.cod +
              rnorm(n * k * 2, 0, 200)
```

### Part 4

```r
gendat_ldt <- function(n = 20, k = 20,
                       b0 = 725, b1 = 50,
                       sigma_u0 = 150, sigma_u1 = 20,
                       sigma = 200) {
  u <- data.frame(subject = 1:n,
                  u0 = rnorm(n, 0, sigma_u0),
                  u1 = rnorm(n, 0, sigma_u1))
  cond.cod <- rep(c(-0.5, 0.5), each = n * k)
  subj     <- rep(rep(1:n, each = k), times = 2)
  sim_dat  <- data.frame(subj = subj, cond.cod = cond.cod)
  nrows <- dim(sim_dat)[1]
  rt    <- rep(NA, nrows)
  for (i in 1:nrows) {
    rt[i] <- b0 +
             u[sim_dat[i, ]$subj, ]$u0 +
             (b1 + u[sim_dat[i, ]$subj, ]$u1) * sim_dat[i, ]$cond.cod +
             rnorm(1, 0, sigma)
  }
  sim_dat$rt <- rt
  sim_dat
}
```

### Part 5

```r
gendat_ldt_corr <- function(n = 20, k = 20,
                            b0 = 725, b1 = 50,
                            sigma_u0 = 150, sigma_u1 = 20,
                            rho = -0.5,            # NEW
                            sigma = 200) {
  Sigma <- matrix(c(sigma_u0^2,                  sigma_u0 * sigma_u1 * rho,
                    sigma_u0 * sigma_u1 * rho,   sigma_u1^2), ncol = 2)
  u <- as.data.frame(mvrnorm(n = n, mu = c(0, 0), Sigma = Sigma))
  colnames(u) <- c("u0", "u1"); u$subject <- 1:n
  cond.cod <- rep(c(-0.5, 0.5), each = n * k)
  subj     <- rep(rep(1:n, each = k), times = 2)
  sim_dat  <- data.frame(subj = subj, cond.cod = cond.cod)
  nrows <- dim(sim_dat)[1]; rt <- rep(NA, nrows)
  for (i in 1:nrows) {
    rt[i] <- b0 + u[sim_dat[i, ]$subj, ]$u0 +
             (b1 + u[sim_dat[i, ]$subj, ]$u1) * sim_dat[i, ]$cond.cod +
             rnorm(1, 0, sigma)
  }
  sim_dat$rt <- rt; sim_dat
}
```

### Part 6

```r
gendat_ldt_crossed <- function(n = 20, k = 20,
                               b0 = 725, b1 = 50,
                               sigma_u0 = 150, sigma_u1 = 20, rho = -0.5,
                               sigma_w = 100,         # NEW: item SD
                               sigma = 200) {
  sim_dat          <- expand.grid(item = 1:(k*2), subj = 1:n)    # fully crossed
  sim_dat$cond.cod <- rep(c(-0.5, 0.5), each = k)
  Sigma_u <- matrix(c(sigma_u0^2,                  rho * sigma_u0 * sigma_u1,
                      rho * sigma_u0 * sigma_u1,   sigma_u1^2), nrow = 2)
  u <- mvrnorm(n = length(unique(sim_dat$subj)), mu = c(0, 0), Sigma = Sigma_u)
  w <- rnorm(n = length(unique(sim_dat$item)), mean = 0, sd = sigma_w)
  N <- dim(sim_dat)[1]; rt <- rep(NA, N)
  for (i in 1:N) {
    rt[i] <- rnorm(1,
                   b0 + u[sim_dat[i, ]$subj, 1] + w[sim_dat[i, ]$item] +
                   (b1 + u[sim_dat[i, ]$subj, 2]) * sim_dat[i, ]$cond.cod,
                   sigma)
  }
  sim_dat$rt <- rt
  sim_dat$subj <- factor(sim_dat$subj); sim_dat$item <- factor(sim_dat$item)
  sim_dat
}
```

### Part 7

```r
gendat_ldt_crossed_log <- function(n = 20, k = 20,
                                   b0 = 6.59,            # log scale
                                   b1 = 0.10,            # log scale
                                   sigma_u0 = 0.2, sigma_u1 = 0.02, rho = -0.5,
                                   sigma_w = 0.1,
                                   sigma = 0.25) {       # all on log scale
  sim_dat          <- expand.grid(item = 1:(k*2), subj = 1:n)
  sim_dat$cond.cod <- rep(c(-0.5, 0.5), each = k)
  Sigma_u <- matrix(c(sigma_u0^2,                  rho * sigma_u0 * sigma_u1,
                      rho * sigma_u0 * sigma_u1,   sigma_u1^2), nrow = 2)
  u <- mvrnorm(n = length(unique(sim_dat$subj)), mu = c(0, 0), Sigma = Sigma_u)
  w <- rnorm(n = length(unique(sim_dat$item)), mean = 0, sd = sigma_w)
  N <- dim(sim_dat)[1]; rt <- rep(NA, N)
  for (i in 1:N) {
    rt[i] <- rlnorm(1,                       # rlnorm() not rnorm()
                    b0 + u[sim_dat[i, ]$subj, 1] + w[sim_dat[i, ]$item] +
                    (b1 + u[sim_dat[i, ]$subj, 2]) * sim_dat[i, ]$cond.cod,
                    sigma)
  }
  sim_dat$rt <- rt
  sim_dat$subj <- factor(sim_dat$subj); sim_dat$item <- factor(sim_dat$item)
  sim_dat
}
```

### Part 8

**Same generator as Part 7. The new code is the nested grid loop:**

```r
# 1D grid (subjects x effect):
for (row in 1:nrow(power_grid)) {
  n_subj <- power_grid$n[row]
  effect <- power_grid$b1[row]
  pvals  <- rep(NA, nsim)
  for (i in 1:nsim) {
    sim_dat <- gendat_ldt_crossed_log(n = n_subj, k = n_item, b1 = effect, ...)
    m <- lmer(log(rt) ~ cond.cod + (1 + cond.cod | subj) + (1 | item), sim_dat)
    pvals[i] <- summary(m)$coefficients[2, 5]
  }
  power_grid$power[row] <- mean(pvals < 0.05)
}
```

:::

## Summary

::: incremental
- Each new ingredient is **additive**: every formula above contains all the previous ones inside.
- The function names track the cumulative complexity: `gendat_ldt` → `_corr` → `_crossed` → `_crossed_log`.
- The corresponding `lmer()` formula also grows: starting from `lm(rt ~ cond.cod)` and ending at `lmer(log(rt) ~ cond.cod + (1 + cond.cod | subj) + (1 | item))`.
:::

## Preguntas? {background-color="#43464B"}

## Session information {.smaller}

```{r}
#| label: session-info
#| echo: false
#| eval: true

sessionInfo()
```
