---
title: "Exercise 1: Power Simulation for Between-Subject Design"
subtitle: "From Part 1: Introduction to simulations and between-subject design"
author:
  - name: "Filip Andras, David López-García & David Sánchez Casasola"
    affiliations:
      - name: "Centro de Investigación Mente Cerebro y Comportamiento (CIMCYC), Universidad de Granada"
date: "26 May 2026"
date-format: "D MMMM YYYY"
format:
  html:
    theme: cosmo
    toc: true
    toc-location: left
    toc-depth: 3
    embed-resources: true
    code-fold: false
    code-overflow: wrap
    code-tools: false
    fig-align: center
    number-sections: true
    page-layout: full
    include-in-header:
      - text: |
          <style>
            pre { white-space: pre-wrap; word-wrap: break-word; }
            .quarto-title-block { max-width: 1200px; margin: auto; padding: 0 2em; }
            main, #quarto-document-content { max-width: 1200px; margin: auto; padding: 0 2em; }
            h2 { margin-top: 2em; }
            h3 { margin-top: 1.5em; }
            #TOC, #quarto-margin-sidebar { width: 250px; }
          </style>
execute:
  warning: false
  message: false
editor: visual
---

# Setup

First, load the required packages:

```{r}
#| label: setup
#| echo: true
#| eval: true

suppressPackageStartupMessages({
  library(tictoc)
})

options(scipen = 999) # prevents scientific notation
options(width = 120)  # wider console output for HTML rendering
set.seed(42)          # set seed for reproducibility
```

# The data generation function

This function generates simulated data for a between-subjects lexical decision
task. We use it throughout the exercise:

```{r}
#| label: define-betwsubj-simdat-ldt
#| echo: true
#| eval: true

betwsubj_simdat_ldt <- function(n = 10,
                                mean_freq_high = 700,
                                mean_freq_low = 750,
                                stdev = 200) {
  # generate data:
  freq_high <- rnorm(n, mean = mean_freq_high, sd = stdev)
  freq_low  <- rnorm(n, mean = mean_freq_low,  sd = stdev)

  subj <- 1:(2 * n)
  cond <- factor(rep(c("freq_high", "freq_low"), each = n))
  cond.cod <- ifelse(cond == "freq_high", -0.5, 0.5)

  # put it together:
  sim_dat <- data.frame(
    subj = subj,
    cond = cond,
    cond.cod = cond.cod,
    rt = c(freq_high, freq_low)
  )
}
```

# Exercise instructions

Your goal: find the sample size per group that gives you **at least 80%
power** for each of the three scenarios below.

The template chunk in each tab below contains a simulation loop. **Change the
value of `n`** (and only `n`) until the printed power is at least **0.80**.
The means and the standard deviation are pre-filled per question.

# Questions to answer

1. **Δ = 50 ms, SD = 200 ms**: how many subjects per group?
2. **Δ = 30 ms, SD = 200 ms** (a smaller effect): how many subjects per group?
3. **Δ = 50 ms, SD = 100 ms** (half the noise): how many subjects per group?

## Number of simulations

Set this once — it is shared by all three questions below:

```{r}
#| label: exercise-nsim
#| echo: true
#| eval: true

nsim <- 1000  # number of simulations per estimate
```

## Your turn

In each tab below, **only the line `n <- 10` is meant to change**. The
design constants (`mean_freq_high`, `mean_freq_low`, `stdev`) are fixed by
the question. The simulation loop is the same template in every tab — leave
it untouched.

::: panel-tabset

### Q1 (Δ=50, SD=200)

```{r}
#| label: exercise-q1-design
#| echo: true
#| eval: false

# Fixed by the question — leave as is:
mean_freq_high <- 700   # high-frequency words
mean_freq_low  <- 750   # low-frequency words (50 ms slower)
stdev          <- 200
```

```{r}
#| label: exercise-q1
#| echo: true
#| eval: false

n <- 10   # <-- CHANGE ONLY THIS until power is at least 0.80

# Simulation loop (same template in every question — leave untouched):
pvals <- rep(NA, nsim)
tic()
for (i in 1:nsim) {
  sim_dat  <- betwsubj_simdat_ldt(n              = n,
                                  mean_freq_high = mean_freq_high,
                                  mean_freq_low  = mean_freq_low,
                                  stdev          = stdev)
  pvals[i] <- t.test(rt ~ cond.cod, data = sim_dat)$p.value
}
toc()

# proportion of significant simulations = estimated power
round(mean(pvals < 0.05), 3)
```

### Q2 (Δ=30, SD=200)

```{r}
#| label: exercise-q2-design
#| echo: true
#| eval: false

# Fixed by the question — leave as is:
mean_freq_high <- 700
mean_freq_low  <- 730   # 30 ms difference (smaller effect)
stdev          <- 200
```

```{r}
#| label: exercise-q2
#| echo: true
#| eval: false

n <- 10   # <-- CHANGE ONLY THIS until power is at least 0.80

# Simulation loop (same template — leave untouched):
pvals <- rep(NA, nsim)
tic()
for (i in 1:nsim) {
  sim_dat  <- betwsubj_simdat_ldt(n              = n,
                                  mean_freq_high = mean_freq_high,
                                  mean_freq_low  = mean_freq_low,
                                  stdev          = stdev)
  pvals[i] <- t.test(rt ~ cond.cod, data = sim_dat)$p.value
}
toc()

round(mean(pvals < 0.05), 3)
```

### Q3 (Δ=50, SD=100)

```{r}
#| label: exercise-q3-design
#| echo: true
#| eval: false

# Fixed by the question — leave as is:
mean_freq_high <- 700
mean_freq_low  <- 750
stdev          <- 100   # half the noise
```

```{r}
#| label: exercise-q3
#| echo: true
#| eval: false

n <- 10   # <-- CHANGE ONLY THIS until power is at least 0.80

# Simulation loop (same template — leave untouched):
pvals <- rep(NA, nsim)
tic()
for (i in 1:nsim) {
  sim_dat  <- betwsubj_simdat_ldt(n              = n,
                                  mean_freq_high = mean_freq_high,
                                  mean_freq_low  = mean_freq_low,
                                  stdev          = stdev)
  pvals[i] <- t.test(rt ~ cond.cod, data = sim_dat)$p.value
}
toc()

round(mean(pvals < 0.05), 3)
```

:::

::: callout-tip
**Strategy.** Start with `n = 10` and roughly double it each time you do not
reach 80% power. Once you overshoot, narrow down. You should converge in
3-4 tries per question.
:::

::: callout-note
The worked solutions live in a separate document:
**`Exercise_1_Power_Simulation_Between_Subjects_solutions.qmd`**. Try the
questions first — your answers will be more useful than reading the
solutions cold.
:::

# Session information

```{r}
#| label: session-info
#| echo: false
#| eval: true
#| code-fold: true

sessionInfo()
```
